---
title: "Week 4: Clustering, cluster stability, and marker genes"
author: "Kevin Z. Lin"
date: "2026-08-15"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE, autodep = TRUE)
knitr::dep_auto()
```

> **A note on `cache = TRUE`, and on the `autodep` line next to it.** This
> document caches every chunk, so the first knit takes tens of minutes and every
> later one takes a couple. Three things are worth knowing before you rely on it.
>
> **Disk.** `knitr` re-saves every object a chunk *modifies*, and a 40,000-cell
> Seurat object is roughly a gigabyte, so `_cache/` reaches several GB. Two
> things keep that in check here. The steps that change the Seurat object are
> grouped into few chunks; and **every chunk whose only output is a figure is
> marked `cache=FALSE`**. That second one is not an aesthetic choice: a `ggplot`
> object stores the environment it was built in, so caching one can drag a copy
> of the entire Seurat object onto disk with it. Plots are cheap to redraw once
> the object is cached upstream, so there is nothing to gain by storing them.
> Also: if your project lives in Dropbox, OneDrive or Google Drive, exclude
> `_cache/` from syncing.
>
> **Interruption.** Stop a knit part-way and the cache can be left truncated,
> which surfaces on the next run as `error reading from connection`. Delete
> `_cache/` and knit again.
>
> **Staleness, which is the one that will actually cost you.** By default
> `knitr` caches each chunk independently and does **not** re-run a chunk when
> the objects it depends on change. Fix a bug upstream, re-knit, and a downstream
> figure can silently stay the *old* figure — a document whose prose and plots
> disagree, with nothing to warn you. `autodep = TRUE` with `knitr::dep_auto()`
> makes `knitr` track which chunks read which objects and invalidate
> accordingly. It is not perfect. Before a document leaves your hands, delete
> `_cache/` and knit once from clean.

# Overview

This tutorial was made from the Seurat guided clustering vignette,
[https://satijalab.org/seurat/articles/pbmc3k_tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial),
and the Seurat differential expression vignette,
[https://satijalab.org/seurat/articles/de_vignette](https://satijalab.org/seurat/articles/de_vignette).
The cluster-comparison, stability and marker-panel sections draw on the
diagnostics used in production snRNA-seq pipelines for human brain tissue.

Week 1 clustered PBMCs on Seurat defaults and did not defend a single choice.
Week 3 normalized properly and did batch correction. This tutorial takes the
clustering step seriously, on a dataset where the defaults do not work:
**40,000 microglial nuclei from 89 human donors.**

Three things make this harder than the vignette:

1. There is no ground truth. In PBMCs you can check your clusters against
   textbook immunology. Microglial states are an open research question, which is
   why people are running these experiments in the first place.
2. **The donors are a potential batch effect.** With 89 of them, an uncorrected
   analysis can produce clusters that are really describing a few individuals.
   Whether it *has* is a question you have to measure, not assume.
3. Cluster boundaries are a modelling output, not an observation. Change the
   resolution and you change how many cell types you "found".

So this tutorial spends most of its time on a question the vignette never asks:
**should you believe these clusters?**

The plan:

| Stage | What it produces | Function |
|---|---|---|
| 1. The naive route | A PCA that has not been batch-corrected | `RunPCA()` |
| 2. Cluster it | Clusters, and a donor-purity problem | `FindClusters()` |
| 3. The corrected route | Clusters from an integrated embedding | `FindClusters()` on `scVI` |
| 4. Do they agree? | A concordance heatmap between the two | `table()`, `pheatmap()` |
| 5. Are they stable? | A per-cluster bootstrap coherence score | `bootstrapStability()` |
| 6. Pick a resolution | Clusters at several resolutions, compared | `FindClusters()` |
| 7. Name them | Marker genes, and known panels | `FindAllMarkers()`, `DotPlot()` |
| 8. Compare to a reference | Our clusters against SEA-AD's labels | `table()` |

> **Where this sits in the workshop.** This is the second half of batch
> correction: Week 3 produced an integrated embedding and stopped; here we use
> one and ask whether it changed the answer. The companion Week 4 tutorial takes
> the labelled cells and tests genes between **donors**. Bulk deconvolution is
> **Week 5**.

# Setting up

```{r install}
# install.packages(c("Seurat", "ggplot2", "pheatmap", "patchwork"))
# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("bluster", "EnhancedVolcano"))
# devtools::install_github("immunogenomics/presto")  # makes wilcox much faster
```

```{r libraries}
library(Seurat)
library(SeuratObject)
library(bluster)
library(EnhancedVolcano)
library(pheatmap)
library(patchwork)
library(ggplot2)
```

```{r clean_env}
rm(list = ls())
```

# The data

**SEA-AD** (Seattle Alzheimer's Disease Brain Cell Atlas) profiled middle
temporal gyrus from an aged cohort spanning the full range of Alzheimer's
neuropathology. This file is the microglia and perivascular-macrophage subset,
downsampled to 40,000 nuclei.

The file downloads itself into a temporary folder, so this tutorial runs as-is
with nothing to set up. It is **525 MB**, so expect the first run to take a few
minutes on a decent connection.

```{r data_dir}
data_dir <- tempdir(check = TRUE)

# If you have already downloaded seaad_microglia.RData, point data_dir at the
# folder holding it instead and the download below will be skipped, e.g.:
# data_dir <- "/path/to/your/data"
```

```{r download}
options(timeout = 3600)

data_url <- paste0("https://www.dropbox.com/scl/fi/cei02rwh3gx61uu9zzeql/",
                   "seaad_microglia.RData",
                   "?rlkey=04erfwn3w2lrnr80pefwfh8fu&dl=1")
data_file <- file.path(data_dir, "seaad_microglia.RData")
expected_size <- 525672784

if (!file.exists(data_file) || file.size(data_file) != expected_size) {
  download.file(data_url, destfile = data_file, mode = "wb")
}

file.size(data_file) == expected_size
```

> **Three details in that download, and every one of them has bitten someone.**
>
> **`options(timeout = 3600)`.** R's default download timeout is **60 seconds**,
> which is not enough for half a gigabyte on any normal connection. Without this
> line the transfer dies partway through with
> `downloaded length 106201070 != reported length 525672784` — and because that
> arrives as a *warning* rather than an error, a script can sail past it and fail
> later at `load()` with a message about a corrupt file. This is the single most
> common reason a large `download.file()` "doesn't work" in R.
>
> **`dl=1`, not `dl=0`.** Copying a Dropbox share link gives you `dl=0`, which
> serves the HTML preview page. You would download a web page, and `load()` would
> complain about a bad magic number.
>
> **`mode = "wb"`.** Forces a binary write. Without it the file is silently
> corrupted on Windows.
>
> The size comparison at the end is the point of all three: **verify the bytes you
> got are the bytes you wanted**, rather than trusting that no error appeared.

```{r load_data}
load(data_file)
seurat_obj
```

Read that summary carefully — it tells you the work someone else already did.

```{r inspect_object}
SeuratObject::Layers(seurat_obj[["RNA"]])
Seurat::Reductions(seurat_obj)
```

Two layers, so the counts have already been log-normalized. Two reductions, and
the interesting one is **`scVI`**: the SEA-AD authors integrated across donors
with a variational autoencoder, and shipped the result. Week 3 named scVI as one
of the methods it was deferring — here it is, pre-computed, which lets us treat
it as a given and spend our time on what it changes.

```{r inspect_reductions}
dim(Seurat::Embeddings(seurat_obj, "scVI"))
dim(Seurat::Embeddings(seurat_obj, "umap"))
```

Ten latent dimensions. The `umap` is the authors' UMAP built from those.

```{r donor_structure}
length(unique(seurat_obj$donor_id))
table(seurat_obj$assay)
table(seurat_obj$Supertype)
```

`Supertype` is SEA-AD's own annotation. We will not look at it again until the
end, where it becomes an independent check on clusters we derived without it.

> **Keeping a label in reserve is a method, not a nicety.** Week 3 used Azimuth
> labels the same way: to judge an embedding you need a read on cell identity
> that was not computed *from* that embedding. Anything you derive from the
> clustering cannot then be used to validate the clustering.

```{r shipped_umap, fig.width=9, fig.height=4.5, cache=FALSE}
gg1 <- Seurat::DimPlot(seurat_obj, reduction = "umap", group.by = "Supertype",
                       raster = TRUE) +
  ggplot2::ggtitle("SEA-AD supertypes")
gg2 <- Seurat::DimPlot(seurat_obj, reduction = "umap", group.by = "donor_id",
                       raster = TRUE) +
  Seurat::NoLegend() +
  ggplot2::ggtitle("Donor (89 of them)")
plot(gg1 + gg2)
```

# The naive route: cluster without correcting for donor

First do what Week 1 taught, unchanged, and see where it fails. The data is
already normalized, so we pick variable genes, scale, and run PCA.

```{r variable_features}
seurat_obj <- Seurat::FindVariableFeatures(seurat_obj,
                                           selection.method = "vst",
                                           nfeatures = 2000,
                                           verbose = FALSE)
seurat_obj <- Seurat::ScaleData(seurat_obj, verbose = FALSE)

set.seed(10)
seurat_obj <- Seurat::RunPCA(seurat_obj,
                             features = Seurat::VariableFeatures(seurat_obj),
                             npcs = 30,
                             verbose = FALSE)

seurat_obj[["RNA"]]$scale.data <- NULL

head(Seurat::VariableFeatures(seurat_obj), 20)
```

> **`ScaleData()` does not scale every gene.** By default it operates on
> `VariableFeatures()` only, which is why this is affordable on 40,000 cells.
> Scaling all 36,412 genes would build a dense matrix of about 11 GB.

> **And `scale.data` is disposable once the PCA exists.** Nothing downstream here
> reads it — `DotPlot()` and `FeaturePlot()` use the `data` layer, and the
> clustering uses the reduction — so we drop it. That is 2,000 × 40,000 doubles,
> about 640 MB, removed from every subsequent cache write. `DoHeatmap()` is the
> one common function that would need it back.

How many components to keep? Week 1 read an elbow off a plot. The same judgment
can be automated by taking the first point where the standard deviation stops
dropping appreciably. Both are shown here — the automated rule is useful
precisely because it makes the choice explicit and reproducible.

```{r elbow, fig.width=7, fig.height=4.5, cache=FALSE}
gg <- Seurat::ElbowPlot(seurat_obj, ndims = 30) +
  ggplot2::ggtitle("Uncorrected PCA")
plot(gg)
```

```{r choose_pcs}
pca_stdev <- Seurat::Stdev(seurat_obj, reduction = "pca")
stdev_drop <- abs(diff(pca_stdev))

num_pcs <- min(which(stdev_drop < 0.1))
num_pcs

round(pca_stdev[1:15], 2)
```

```{r cluster_pca}
set.seed(10)
seurat_obj <- Seurat::FindNeighbors(seurat_obj,
                                    reduction = "pca",
                                    dims = 1:num_pcs,
                                    verbose = FALSE)
seurat_obj <- Seurat::FindClusters(seurat_obj,
                                   resolution = 0.5,
                                   cluster.name = "pca_clusters",
                                   verbose = FALSE)

set.seed(10)
seurat_obj <- Seurat::RunUMAP(seurat_obj,
                              reduction = "pca",
                              dims = 1:num_pcs,
                              reduction.name = "umap_pca",
                              verbose = FALSE)

table(seurat_obj$pca_clusters)
```

```{r plot_umap_pca, fig.width=10, fig.height=4.5, cache=FALSE}
gg1 <- Seurat::DimPlot(seurat_obj, reduction = "umap_pca",
                       group.by = "pca_clusters", label = TRUE,
                       raster = TRUE) +
  Seurat::NoLegend() +
  ggplot2::ggtitle("Uncorrected: clusters")
gg2 <- Seurat::DimPlot(seurat_obj, reduction = "umap_pca",
                       group.by = "donor_id", raster = TRUE) +
  Seurat::NoLegend() +
  ggplot2::ggtitle("Uncorrected: donor")
plot(gg1 + gg2)
```

## Measuring the damage instead of eyeballing it

A UMAP coloured by 89 donors is unreadable, and "it looks mixed" is not a result.
Quantify it: for each cluster, what fraction of its cells come from its single
most-represented donor, and how many donors does it take to make up half the
cluster?

```{r donor_purity}
donor_purity <- function(clusters, donors) {
  cluster_levels <- sort(unique(as.character(clusters)))
  out_df <- data.frame(cluster = cluster_levels,
                       n_cells = NA_integer_,
                       top_donor_frac = NA_real_,
                       donors_for_half = NA_integer_)

  for (i in seq_along(cluster_levels)) {
    in_cluster <- as.character(clusters) == cluster_levels[i]
    donor_tab <- sort(table(donors[in_cluster]), decreasing = TRUE)
    out_df[i, "n_cells"] <- sum(in_cluster)
    out_df[i, "top_donor_frac"] <- max(donor_tab) / sum(donor_tab)
    out_df[i, "donors_for_half"] <-
      min(which(cumsum(donor_tab) >= 0.5 * sum(donor_tab)))
  }
  out_df
}

purity_pca <- donor_purity(seurat_obj$pca_clusters, seurat_obj$donor_id)
purity_pca
```

With 89 donors contributing unevenly, a cluster that reflects biology should draw
from many of them. A cluster where one donor supplies a large share, or where two
or three donors supply half, is describing a person rather than a cell state.

# The corrected route: cluster on the integrated embedding

Same graph-based clustering, same resolution — the only change is which
reduction the neighbour graph is built on.

```{r cluster_scvi}
set.seed(10)
seurat_obj <- Seurat::FindNeighbors(seurat_obj,
                                    reduction = "scVI",
                                    dims = 1:10,
                                    verbose = FALSE)
seurat_obj <- Seurat::FindClusters(seurat_obj,
                                   resolution = 0.5,
                                   cluster.name = "scvi_clusters",
                                   verbose = FALSE)

table(seurat_obj$scvi_clusters)
```

```{r plot_umap_scvi, fig.width=10, fig.height=4.5, cache=FALSE}
gg1 <- Seurat::DimPlot(seurat_obj, reduction = "umap",
                       group.by = "scvi_clusters", label = TRUE,
                       raster = TRUE) +
  Seurat::NoLegend() +
  ggplot2::ggtitle("Integrated: clusters")
gg2 <- Seurat::DimPlot(seurat_obj, reduction = "umap",
                       group.by = "donor_id", raster = TRUE) +
  Seurat::NoLegend() +
  ggplot2::ggtitle("Integrated: donor")
plot(gg1 + gg2)
```

```{r donor_purity_scvi}
purity_scvi <- donor_purity(seurat_obj$scvi_clusters, seurat_obj$donor_id)
purity_scvi
```

```{r purity_comparison}
summary(purity_pca[, "top_donor_frac"])
summary(purity_scvi[, "top_donor_frac"])

sum(purity_pca[, "donors_for_half"] <= 3)
sum(purity_scvi[, "donors_for_half"] <= 3)
```

**Integration helped, and by less than the introduction implied.** Uncorrected,
the median cluster draws 10.4% of its cells from its top donor, the worst draws
32.8%, and **one** cluster gets half its cells from three donors or fewer. After
integration the median falls to 7.5%, the worst to 20.8%, and **no** cluster is
that concentrated.

# Do the two clusterings agree?

Purity says the corrected clustering is less donor-driven. It does not say the
two analyses found different *biology*. Cross-tabulate them.

```{r concordance, fig.width=7.5, fig.height=6, cache=FALSE}
concordance_tab <- table(paste0("pca_", seurat_obj$pca_clusters),
                         paste0("scvi_", seurat_obj$scvi_clusters))
concordance_tab

pheatmap::pheatmap(log10(concordance_tab + 10),
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "Uncorrected vs integrated clusters (log10 counts)")
```

The `log10(tab + 10)` is a display trick worth keeping: without the `+ 10`, the
many zero cells dominate the colour scale and the structure disappears.

**How to read it.** One bright cell per row and per column — a near-permutation
matrix — means the two methods recovered the same partition under different
names. A row that spreads across several columns means that cluster was split by
the other method; a column collecting several rows means it was merged.

# Are the clusters stable?

Agreement between two methods is one kind of evidence. Another is whether a
single method finds the same clusters again when you perturb the data.
`bluster::bootstrapStability()` resamples the cells, re-clusters, and reports
how consistently each original cluster is recovered.

```{r stability_subsample}
set.seed(10)
stability_cells <- sample(colnames(seurat_obj), 10000)
scvi_embedding <- Seurat::Embeddings(seurat_obj, "scVI")[stability_cells, ]
dim(scvi_embedding)
```

We subsample to 10,000 cells because this is the one genuinely expensive step —
it re-clusters the data twenty times. That is a compute compromise, and worth
stating rather than hiding.

```{r stability}
cluster_fun <- function(x) {
  graph_obj <- bluster::makeSNNGraph(x, type = "jaccard")
  igraph::cluster_louvain(graph_obj)$membership
}

set.seed(10)
original_clusters <- cluster_fun(scvi_embedding)
table(original_clusters)

set.seed(10)
stability_mat <- bluster::bootstrapStability(scvi_embedding,
                                             FUN = cluster_fun,
                                             clusters = original_clusters)
dim(stability_mat)
round(diag(stability_mat), 2)
```

```{r stability_plot, fig.width=6.5, fig.height=5.5, cache=FALSE}
pheatmap::pheatmap(stability_mat,
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "Bootstrap cluster coherence")
```

The plot is **upper-triangular** — `pairwiseRand()` fills only one half, so the
grey lower block is structurally empty rather than missing data. Read the
diagonal for how coherent each cluster is, and the off-diagonal for whether a
given *pair* stays separated. Here most off-diagonal cells are red (pairs do stay
apart) while most diagonal cells are pale (individual clusters do not hold
together) — which is the signature of a set of boundaries drawn across a
continuum.

The `bluster` documentation states the reading of this plot exactly right, so
here it is verbatim:

> High on-diagonal values indicate that the corresponding cluster remains
> coherent in the bootstrap replicates, while high off-diagonal values indicate
> that the corresponding pair of clusters are still separated in the replicates.

A low diagonal entry is a cluster that dissolves when the data is perturbed —
which does not mean it is fake, but does mean you should not build a story on it
without other evidence.

**And this is the least comfortable result in the tutorial.** Of the 14 clusters
found on the subsample, three come back with coherence above 0.9 (0.98, 0.96,
0.93) and one at 0.54. **The other ten sit between 0.12 and 0.26.** Resample the
cells and most of these clusters do not reappear as the same groups.

Take that seriously rather than explaining it away. Microglia in cortex are not
a set of cleanly separated cell types the way lymphocytes in blood are; much of
the variation is a continuum of activation state, and cutting a continuum into
communities is exactly the situation where the boundaries move under
resampling. Three of these clusters look like genuinely distinct populations.
The rest are better described as regions of a gradient than as types — and a
Louvain algorithm will return them with the same confident integer labels
either way.

> **What to do with an unstable cluster.** Not necessarily discard it. But do not
> name it as a cell type, do not build a downstream comparison that depends on
> its exact boundary, and do not report the number of clusters as a finding. If a
> result survives only at one resolution on one clustering, it is a property of
> the algorithm.

# Choosing a resolution

`resolution` is the parameter with the largest effect on your conclusions and the
least justification in most papers. There is no correct value; there is a value
you can defend.

```{r resolutions}
set.seed(10)
seurat_obj <- Seurat::FindClusters(seurat_obj,
                                   resolution = c(0.1, 0.2, 0.5, 1.0),
                                   verbose = FALSE)

resolution_cols <- c("RNA_snn_res.0.1", "RNA_snn_res.0.2",
                     "RNA_snn_res.0.5", "RNA_snn_res.1")
sapply(resolution_cols,
       function(x) length(unique(seurat_obj@meta.data[, x])))
```

```{r resolution_plots, fig.width=11, fig.height=9, cache=FALSE}
gg_list <- lapply(resolution_cols, function(x) {
  Seurat::DimPlot(seurat_obj, reduction = "umap", group.by = x,
                  label = TRUE, raster = TRUE) +
    Seurat::NoLegend() +
    ggplot2::ggtitle(x)
})
plot(patchwork::wrap_plots(gg_list, ncol = 2))
```

## Where are the cells actually dense?

Before either check, it is worth looking at the UMAP without any cluster colours
on it. A `DimPlot` coloured by cluster shows you the partition you asked for, and
a partition always looks like a partition — the colours tell you where the
boundaries were drawn, not whether there was anything there to divide. Plotting
the **density** of cells instead answers a different question: how many
well-separated lumps does this embedding actually contain?

```{r density_umap, fig.width=11, fig.height=4.5, cache=FALSE}
umap_df <- Seurat::FetchData(seurat_obj, vars = c("umap_1", "umap_2"))

gg1 <- ggplot2::ggplot(umap_df, ggplot2::aes(x = umap_1, y = umap_2)) +
  ggplot2::geom_point(alpha = 0.15, size = 0.01) +
  ggplot2::geom_density2d(colour = "red", bins = 12) +
  ggplot2::ggtitle("Cell density, contour lines") +
  ggplot2::theme_bw()

gg2 <- ggplot2::ggplot(umap_df, ggplot2::aes(x = umap_1, y = umap_2)) +
  ggplot2::stat_density_2d(geom = "polygon", contour = TRUE,
                           ggplot2::aes(fill = ggplot2::after_stat(level)),
                           bins = 12, alpha = 0.4) +
  ggplot2::scale_fill_distiller(palette = "Blues", direction = 1) +
  ggplot2::geom_density2d(colour = "black", bins = 12, linewidth = 0.2) +
  ggplot2::ggtitle("Cell density, filled contours") +
  ggplot2::theme_bw()

plot(gg1 + gg2)
```

Count the separated masses, then count the peaks inside them. This embedding has
**three small, cleanly detached islands** — far left, bottom centre, and a tiny
one at the left edge of the main body — plus **one large mass containing about
two internal peaks**. Call it four or five modes.

Now compare that with what we asked for: 6 clusters at resolution 0.2, and 17 at
resolution 1.0. The three islands and the couple of peaks account for maybe five;
every boundary beyond that is being drawn *through* a dense region rather than
between two of them. That is the same conclusion the stability analysis reached,
visible directly and without any resampling — and it lines up with the marker
panels below, where the three islands turn out to be T cells, perivascular
macrophages and a contamination cluster, while the big mass is the microglia.

The same plot split by donor is the visual counterpart to the purity table.

```{r density_by_donor, fig.width=11, fig.height=5, cache=FALSE}
set.seed(10)
donor_subset <- sample(unique(seurat_obj$donor_id), 6)

donor_df <- Seurat::FetchData(seurat_obj,
                              vars = c("umap_1", "umap_2", "donor_id"))
donor_df <- donor_df[donor_df$donor_id %in% donor_subset, ]

gg <- ggplot2::ggplot(donor_df, ggplot2::aes(x = umap_1, y = umap_2)) +
  ggplot2::geom_point(data = umap_df, colour = "grey85", size = 0.01) +
  ggplot2::geom_density2d(colour = "red", bins = 8) +
  ggplot2::facet_wrap(~ donor_id, ncol = 3) +
  ggplot2::ggtitle("Density per donor, six donors, whole dataset in grey") +
  ggplot2::theme_bw()
plot(gg)
```

No donor owns an island — which is the reassuring half, and matches the purity
numbers. But look at *where inside the big mass* each donor peaks: six donors,
six different modal positions, spread across the whole blob. The donor effect
here is not "this person got their own cluster"; it is "this person sits slightly
differently along a continuum". That is a much harder thing to see in a
cluster-coloured UMAP, and it is a good part of why boundaries drawn inside that
mass move when you resample.

> **Read the peaks, not the absences.** A donor with few cells may produce no
> contour at all in a region where it does have cells — contouring needs density,
> so an empty facet region means "not enough cells to form a contour here", not
> "no cells here". Compare where the contours *are*, not where they are missing.

Two checks are more informative than staring at the panels.

**First: does raising the resolution split clusters into donor-specific pieces?**
If new clusters appear that are dominated by a handful of donors, the resolution
has passed the point where it is describing biology.

```{r resolution_purity}
purity_by_resolution <- lapply(resolution_cols, function(x) {
  purity_df <- donor_purity(seurat_obj@meta.data[, x], seurat_obj$donor_id)
  data.frame(resolution = x,
             n_clusters = nrow(purity_df),
             median_top_donor = median(purity_df[, "top_donor_frac"]),
             max_top_donor = max(purity_df[, "top_donor_frac"]),
             min_cluster_size = min(purity_df[, "n_cells"]))
})
do.call(rbind, purity_by_resolution)
```

Here the answer is **no** — and that is worth seeing, because the check is
usually taught as if it always catches something. Across resolutions 0.1 to 1.0
the cluster count goes 4, 6, 9, 17 while the median top-donor share stays
between 7.4% and 9.0% and the worst rises only from 20.5% to 25.0%. The smallest
cluster never falls below 227 cells. On this dataset, turning the resolution up
subdivides real structure rather than carving out individuals. On a dataset with
fewer donors or stronger batch effects it will not, which is why you run it.

**Second: are the smallest clusters big enough to survive the next analysis?**
The companion tutorial sums counts within one cell type per donor, and runs the
same feasibility check before choosing which type to test. A cluster of 200 cells
spread over 89 donors gives about two cells per donor — no donor-level comparison
is possible, and the DE result for that cluster would be noise wearing a p-value.

```{r cells_per_donor_per_cluster}
cells_tab <- table(seurat_obj$RNA_snn_res.0.2, seurat_obj$donor_id)
apply(cells_tab, 1,
      function(x) c(donors_with_10_or_more = sum(x >= 10),
                    median_cells = median(x)))
```

This is the check that bites. At resolution 0.2, clusters 0 and 1 appear in 87
and 83 donors with at least ten cells (medians of 234 and 107 cells per donor).
Clusters 2 and 3 manage 39 and 47 donors at a median of 9 and 10 cells. Clusters
4 and 5 reach 8 and 3 donors at a median of 1 and 2 cells.

So only two of these six clusters can support a donor-level comparison at all.
The companion tutorial reaches the same conclusion from the other direction, and
it is why it tests a single well-populated cell type rather than looping over
everything.

> **The resolution is a downstream decision, not a visual one.** Pick it knowing
> what you intend to do next. If the plan is donor-level differential expression,
> the binding constraint is cells per donor per cluster, and that constraint is
> checkable before you commit.

We take **resolution 0.2** for the rest of this tutorial.

```{r set_idents}
Seurat::Idents(seurat_obj) <- "RNA_snn_res.0.2"
table(Seurat::Idents(seurat_obj))
```

# Comparing two clusters

The first differential expression question of the session is the *easy* one:
what distinguishes one cluster of cells from another? Here cells genuinely are
the units of analysis — we are describing a partition of cells, not generalizing
to new donors — so a cell-level test is the right tool.

```{r de_two_clusters}
de_res <- Seurat::FindMarkers(seurat_obj,
                              ident.1 = "0",
                              ident.2 = "1",
                              test.use = "wilcox")
head(de_res, 10)
nrow(de_res)
```

`EnhancedVolcano` draws its horizontal line at a **raw** p-value, so passing
`pCutoff = 0.05` would draw the line in the wrong place — that is an uncorrected
threshold on 12,982 simultaneous tests. Convert the multiple-testing threshold
you actually want into the raw p-value that corresponds to it, and pass that.

```{r bh_threshold}
bh_adj <- p.adjust(de_res$p_val, method = "BH")
sum(bh_adj < 0.05)

p_threshold <- max(de_res$p_val[bh_adj < 0.05])
p_threshold
```

Benjamini–Hochberg works by comparing the sorted p-values against a rising line,
so "BH-significant at 0.05" corresponds to a single raw-p boundary — the largest
raw p-value that still made the cut. That number is what the volcano's line
should sit at.

**And look at where it landed: 0.038, with 9,907 of the 12,982 tested genes
passing.** BH barely moved the threshold off the uncorrected 0.05, because the
correction only has work to do when most tests are null — and here three quarters
of them are "significant". That is not evidence of enormous biological signal
between two clusters of the same cell type. It is what the circularity above
looks like when you put a number on it: the test cannot discriminate, so the
multiple-testing correction has nothing to correct.

> **A useful diagnostic in general.** If a multiple-testing procedure leaves your
> threshold essentially where it started, ask why so few of your tests are null.
> Sometimes the answer is a genuinely strong effect. More often it means the null
> hypothesis being tested is not the one you think — as here, where the cells were
> selected for being different before they were tested for being different.

> **Note this is not the column Seurat gives you.** `p_val_adj` in a
> `FindMarkers()` table is a **Bonferroni** correction over every gene in the
> assay, which is far more conservative than BH and is computed over a different
> denominator than the genes actually tested here. We compute BH ourselves over
> the tested set rather than reusing that column.

```{r volcano_clusters, fig.width=9, fig.height=7, cache=FALSE}
gg <- EnhancedVolcano::EnhancedVolcano(
  de_res,
  lab = rownames(de_res),
  x = "avg_log2FC",
  y = "p_val",
  pCutoff = p_threshold,
  FCcutoff = 0.5,
  title = "Cluster 0 versus cluster 1",
  subtitle = "Wilcoxon, cells as units; line at BH-adjusted p < 0.05")
plot(gg)
```

> **Do not read these p-values as evidence.** The clusters were *defined* by
> finding the genes that separate these cells, and we are now testing those same
> cells for those same genes. The p-values are circular by construction, and the
> top ten come back as **exactly zero** — underflowing double precision. Use this
> ranking to **describe** clusters. Never quote it as proof that a cluster is
> real; the stability analysis above is for that, and it was a good deal less
> flattering.

Worth noticing what the top of that list actually contains: `ADAM7-AS1`,
`DLEU7`, `ABHD15-AS1` and several bare `ENSG` identifiers — antisense
transcripts, lncRNAs and unannotated features rather than recognizable marker
genes. All ten are *lower* in cluster 0. A separation carried by unannotated
features is not automatically wrong, but it is the signature of a technical axis
(detection depth, annotation coverage) more often than a biological one, and it
is a reason to look at cluster 0 versus 1 with suspicion rather than to start
writing about `OPRM1` in microglia.

# Marker genes

For naming clusters, what you want is genes that are up in one cluster against
*all* others, not against one chosen rival.

```{r find_all_markers}
all_markers <- Seurat::FindAllMarkers(seurat_obj,
                                      only.pos = TRUE,
                                      test.use = "wilcox",
                                      min.pct = 0.25,
                                      logfc.threshold = 0.25,
                                      verbose = FALSE)
dim(all_markers)
table(all_markers$cluster)
```

```{r top_markers}
top_markers <- do.call(rbind, lapply(split(all_markers, all_markers$cluster),
                                     function(x) {
  x <- x[order(x$avg_log2FC, decreasing = TRUE), ]
  head(x, 5)
}))
top_markers[, c("cluster", "gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")]
```

```{r dotplot_markers, fig.width=11, fig.height=5, cache=FALSE}
marker_genes <- unique(top_markers[, "gene"])
gg <- Seurat::DotPlot(seurat_obj, features = marker_genes) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                                     vjust = 0.5)) +
  ggplot2::ggtitle("Top 5 markers per cluster by fold change")
plot(gg)
```

## Ranking by fold change is not the same as ranking by p-value

```{r ranking_comparison}
cluster0_markers <- all_markers[all_markers$cluster == "0", ]

by_pval <- head(cluster0_markers[order(cluster0_markers$p_val), "gene"], 10)
by_fc <- head(cluster0_markers[order(cluster0_markers$avg_log2FC,
                                     decreasing = TRUE), "gene"], 10)

by_pval
by_fc
length(intersect(by_pval, by_fc))
```

Sorting by p-value favours genes expressed in many cells, because that is what
makes a Wilcoxon test confident. Sorting by fold change favours genes with big
differences, which are often expressed in few cells. Neither is wrong.

**For cluster 0 the two top-ten lists share exactly two genes.** Eight of the ten
markers you would put on a slide depend on which column you sorted by — a choice
that is almost never stated in a figure legend, and one Seurat makes for you if
you do not.

## A curated panel beats a derived list for naming

Data-driven markers tell you what separates your clusters. They do not tell you
what the clusters *are*. For that you need genes whose meaning was established
elsewhere. Production pipelines for human brain tissue typically carry a large
library of such panels, one per expected cell type; this is the microglial
subset of one.

```{r marker_panels}
panel_list <- list(
  homeostatic = c("P2RY12", "P2RY13", "CX3CR1", "TMEM119", "CSF1R", "SALL1"),
  activated = c("APOE", "SPP1", "GPNMB", "LPL", "ITGAX", "CD9"),
  antigen = c("CD74", "HLA-DRA", "HLA-DRB1"),
  pvm = c("CD163", "MRC1", "F13A1", "LYVE1"),
  lymphocyte = c("CD2", "CD3E", "IL7R", "SKAP1", "THEMIS", "CCL5"),
  proliferating = c("MKI67", "TOP2A"),
  other_lineage = c("SNAP25", "PLP1", "SLC1A2", "CLDN5")
)

panel_genes <- unlist(panel_list)
setdiff(panel_genes, rownames(seurat_obj))
panel_genes <- intersect(panel_genes, rownames(seurat_obj))
```

> **Check that your genes exist before plotting them.** `setdiff()` against
> `rownames()` takes one line, and skipping it is how you end up presenting a
> panel with silent holes. It is worth wrapping in a helper if you plot panels
> often — a missing gene should degrade the figure, not error out or
> disappear. Here everything in the panel is present, which is the boring
> outcome and the one you want; on a mouse dataset, or an older annotation, or a
> targeted panel like the spatial data from Week 1, it will not be.

```{r dotplot_panel, fig.width=11, fig.height=5, cache=FALSE}
gg <- Seurat::DotPlot(seurat_obj, features = panel_genes) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1,
                                                     vjust = 0.5)) +
  ggplot2::ggtitle("Curated microglial panels")
plot(gg)
```

```{r feature_plots, fig.width=11, fig.height=7, cache=FALSE}
gg <- Seurat::FeaturePlot(seurat_obj, reduction = "umap",
                          features = c("P2RY12", "CX3CR1", "APOE", "SPP1",
                                       "CD163", "MRC1"),
                          raster = TRUE, ncol = 3)
plot(gg)
```

**This one figure labels the whole dataset, and it is worth reading cluster by
cluster.** Each panel lights up in exactly one place:

| Cluster | Cells | What the panel says |
|---|---|---|
| 0 | 25,818 | Homeostatic microglia — `P2RY12`, `CX3CR1`, `CSF1R`, moderate `APOE`/`SPP1` |
| 1 | 11,507 | Microglia again, but with the antigen-presentation row (`CD74`, `HLA-DRA`, `HLA-DRB1`) markedly higher |
| 2 | 1,119 | **T lymphocytes** — the entire `lymphocyte` row, and nothing else |
| 3 | 1,038 | **Perivascular macrophages** — `CD163`, `MRC1`, `F13A1`, `LYVE1`, all four |
| 4 | 276 | **Proliferating cells** — `MKI67` and `TOP2A`, uniquely |
| 5 | 242 | **Not microglia** — `SNAP25`, `PLP1`, `SLC1A2`: neuronal, oligodendrocyte and astrocyte markers together |

That is a clean answer, and note how little of it came from the differential
expression. The clusters were found without any of these genes; the panel simply
read off what they are. Six clusters, six interpretable identities, from one
dot plot.

Two of those rows deserve emphasis. **Cluster 5 expresses neuronal, oligodendrocyte
*and* astrocyte markers simultaneously** — no real cell does that, so this is
either doublets that survived Week 2's filtering or nuclei carrying heavy ambient
RNA. Yet SEA-AD labels 86% of it `Micro-PVM_2_3`. A published atlas annotation
does not exempt you from looking. And **cluster 3 is perivascular macrophages**,
which is a genuinely different cell type from microglia that the dataset's name
(`Microglia-PVM`) tells you to expect — if you had treated all 40,000 nuclei as
one population, you would have averaged the two together.

The last two rows are quality-control panels, not biology panels. `SNAP25` is
neuronal, `PLP1` oligodendrocyte, `SLC1A2` astrocyte, `CLDN5` endothelial. In a
dataset subset to microglia, any cluster showing these is either a doublet
population that survived Week 2's filtering or ambient RNA from the surrounding
tissue. Either way it is a finding about your data, and the companion tutorial
shows how far that kind of contamination can propagate into a DE result.

> **The `lymphocyte` row is in this panel because the data put it there, and it
> is worth saying how.** The first draft of this tutorial had no lymphocyte
> markers — the QC row went straight from microglia to neurons, oligodendrocytes,
> astrocytes and endothelium, which is the standard brain contamination list.
> Then `FindAllMarkers()` returned `CD2`, `CD3E` and `TRGC2` as the top markers of
> cluster 2, at log2 fold changes above 10 with the genes detected in half that
> cluster and essentially none of the rest. Those are T-cell receptor and T-cell
> co-receptor genes. There is a lymphocyte population sitting inside a dataset
> labelled microglia-and-perivascular-macrophage — brain does contain resident and
> infiltrating T cells, so this is plausible biology rather than an error, but it
> is not microglia.
>
> **The curated panel would have missed it and the derived markers caught it.**
> That is the honest relationship between the two approaches: a panel encodes what
> you already expected to find, so it can only confirm or deny those things.
> `FindAllMarkers()` has no expectations. Run both, and when the data-driven list
> names something your panel does not cover, extend the panel — which is what the
> `lymphocyte` row is.

# Checking against a reference annotation

Now bring back the label we set aside. SEA-AD's `Supertype` was assigned by the
atlas authors using far more data than we have here, so it is not truth — but it
is independent of everything we just did.

```{r compare_supertype, fig.width=8, fig.height=5, cache=FALSE}
supertype_tab <- table(paste0("cluster_", Seurat::Idents(seurat_obj)),
                       seurat_obj$Supertype)
supertype_tab

pheatmap::pheatmap(log10(supertype_tab + 10),
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "Our clusters vs SEA-AD supertypes (log10 counts)")
```

```{r supertype_proportions}
round(prop.table(supertype_tab, margin = 1), 2)
```

Rows are our clusters; each row shows how that cluster's cells were labelled by
SEA-AD. A cluster that maps cleanly onto one supertype has independent support.
A cluster that splits a supertype may be finer structure, or may be noise.

The result splits neatly in two. Four of our six clusters map onto a single
supertype at 73% or better — cluster 3 is 98% `Micro-PVM_1`, cluster 5 is 86%
`Micro-PVM_2_3`, cluster 1 is 84% `Micro-PVM_2`, cluster 2 is 74%
`Micro-PVM_2_2`. Those have support from an annotation that never saw our
clustering. Cluster 0, the largest at 25,818 cells, does not: it is 49%
`Micro-PVM_2`, 31% `Micro-PVM_2_3` and 16% `Micro-PVM_3`, a blend of three
supertypes that SEA-AD considered distinct.

Line that up with the marker panels and with the stability result, and all three
tell one story. The clusters with a clean identity — T cells, perivascular
macrophages, proliferating cells, the contamination cluster — are small,
distinct, recognized by the reference, and (in the stability analysis) the kind
that reproduce under resampling. The large microglial clusters 0 and 1 are the
ones that blur into each other, spread across supertypes, and dissolve when the
data is perturbed.

That is not a failure of the analysis. It is the actual structure of the data:
**a few genuinely separate cell types, plus a large microglial continuum that
clustering reports as though it were made of types.** Knowing which of your
clusters is which kind is the difference between a defensible figure and an
overinterpreted one — and it took three independent checks, none of which was a
p-value.

> **Disagreement is not automatically your error.** A reference built on a
> different cohort, at a different resolution, with a different objective, will
> disagree in places. The value of the comparison is that it tells you *where* to
> look, not who is right.

# What we glossed over

- **We used someone else's integration.** The `scVI` embedding came with the
  file. Running scVI yourself needs Python and `scvi-tools`; Harmony and
  `IntegrateLayers()` are the in-R alternatives shown in **Week 3**. Comparing
  integration methods head to head — with the purity and concordance measures
  used above — is the natural next step and was not done here.
- **Clustering algorithm.** We used Louvain, Seurat's default. Leiden
  (`algorithm = 4`) is generally better behaved and guarantees connected
  communities. Running both and comparing them with the concordance heatmap used
  above is a cheap check that your clusters are not an artifact of one algorithm.
- **Number of PCs, for the uncorrected route only.** The integrated route used
  all 10 scVI dimensions because that is all there are. The choice matters more
  than the resolution in some datasets and we changed only one of them at a time.
- **Doublets and ambient RNA.** This object arrives pre-cleaned by the SEA-AD
  authors, so **Week 2**'s material is invisible here. Your own data will not
  arrive that way, and the `other_lineage` panel above is how you check.
- **Composition testing.** We counted cells per cluster per donor to check
  feasibility. Asking whether those *proportions* differ between disease groups
  is a genuine analysis with its own methods (`scCODA`, `propeller`), and it is
  not the same thing as differential expression.
- **Automated cell-type labelling.** MapMyCells and Azimuth assign labels against
  a reference directly, rather than clustering first and naming afterwards.
  Week 3 used Azimuth on blood; the brain references are the relevant ones here.
- **Cluster-level multiple testing.** `FindAllMarkers()` computes a Bonferroni
  adjustment within each cluster, over all genes. It does not adjust across
  clusters, and the clusters were chosen from the same data.

# References

- Hao Y *et al.* (2024). "Dictionary learning for integrative, multimodal and
  scalable single-cell analysis" (Seurat v5). *Nature Biotechnology* 42:293–304.
  [doi:10.1038/s41587-023-01767-y](https://doi.org/10.1038/s41587-023-01767-y)
- Lun ATL, Risso D. `bluster`: clustering algorithms and cluster diagnostics.
  [https://bioconductor.org/packages/bluster](https://bioconductor.org/packages/bluster)
- Traag VA, Waltman L, van Eck NJ (2019). "From Louvain to Leiden: guaranteeing
  well-connected communities." *Scientific Reports* 9:5233.
  [doi:10.1038/s41598-019-41695-z](https://doi.org/10.1038/s41598-019-41695-z)
- Lopez R *et al.* (2018). "Deep generative modeling for single-cell
  transcriptomics" (scVI). *Nature Methods* 15:1053–1058.
  [doi:10.1038/s41592-018-0229-2](https://doi.org/10.1038/s41592-018-0229-2)
- Prater KE, Lin KZ (2025). "All the single cells: single-cell
  transcriptomics/epigenomics experimental design and analysis considerations for
  glial biologists." *Glia* 73:451–473.
  [doi:10.1002/glia.24633](https://doi.org/10.1002/glia.24633)
  — §3.5 on cluster stability and §3.6 on cell type annotation cover this
  tutorial directly.
- Gabitto MI *et al.* (2024). "Integrated multimodal cell atlas of Alzheimer's
  disease" (SEA-AD). *Nature Neuroscience*.
  [https://portal.brain-map.org/explore/seattle-alzheimers-disease](https://portal.brain-map.org/explore/seattle-alzheimers-disease)
