---
title: "Week 4: Batch correction, part 2 — integrating two studies, and correcting the expression matrix"
author: "Kevin Z. Lin"
date: "2026-08-17"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE, autodep = TRUE)
knitr::dep_auto()
```

> **A note on `cache = TRUE` and `autodep`.** As with the other two Week 4
> documents, every chunk is cached, so the first knit takes a long time and every
> later one takes a couple of minutes. Chunks whose only output is a figure are
> marked `cache=FALSE`, because a `ggplot` object stores the environment it was
> built in and caching one can drag a copy of the whole Seurat object onto disk.
> `autodep = TRUE` plus `knitr::dep_auto()` makes `knitr` re-run a chunk when the
> objects it reads have changed — without it a fixed bug upstream can leave a
> stale figure downstream, with nothing to warn you. Delete `_cache/` and knit
> once from clean before the document leaves your hands, and exclude `_cache/`
> from Dropbox or OneDrive syncing.

# Overview

This tutorial was made from the Seurat v4 vignette *Introduction to scRNA-seq
integration*,
[https://satijalab.org/seurat/archive/v4.3/integration_introduction](https://satijalab.org/seurat/archive/v4.3/integration_introduction),
and *Mapping and annotating query datasets*,
[https://satijalab.org/seurat/articles/integration_mapping](https://satijalab.org/seurat/articles/integration_mapping).

**Week 3 stopped one step short on purpose.** It ran `IntegrateLayers()`, got a
new *reduction* — a set of coordinates with no gene names attached — and said, in
so many words: some methods can also give you a corrected expression matrix, it
is tempting to test genes on it, do not, and Week 4 will explain why. This is
that explanation, and it starts by actually building one.

Two things are different here.

**The batch is a whole study.** Week 3's batches were nine protocols run on the
same PBMCs. Here the two batches are two independent research programs — SEA-AD
in Seattle and ROSMAP through MIT and Rush — with different dissections,
different chemistries, different reference annotations, different QC decisions,
and no shared donors at all. This is the hard case, and it is also the common
one: it is what happens the moment you want to say anything about your data in
the context of somebody else's.

**The output is an expression matrix, not just an embedding.**
`FindIntegrationAnchors()` followed by `IntegrateData()` returns an object with a
second **assay** in it, whose values are genes × cells and which you can put
straight into `FeaturePlot()`, `DotPlot()`, or — and this is the whole point of
the tutorial — into `FindMarkers()`, where it will cheerfully give you p-values
that mean nothing.

The plan:

| Stage | What it produces | Function |
|---|---|---|
| 1. Make the two objects comparable | Shared barcodes, shared genes, comparable sizes | `RenameCells()`, `subset()` |
| 2. Look at them uncorrected | A UMAP with one island per study | `merge()`, `RunPCA()`, `RunUMAP()` |
| 3. Measure the separation | A per-cell batch-mixing score, against its null | `RANN::nn2()` |
| 4. Find anchors | Pairs of cells the method believes are the same cell | `FindIntegrationAnchors()` |
| 5. Correct the expression | A new **assay**, `integrated`, of corrected values | `IntegrateData()` |
| 6. Look inside it | Fewer genes, no counts, and negative numbers | `LayerData()` |
| 7. Re-embed, and re-measure | The same mixing score, on corrected values | `RunPCA()`, `RunUMAP()` |
| 8. Go back to the counts | The assay you actually test on | `JoinLayers()` |
| 9. The alternative | Reference labels moved onto the query, counts untouched | `TransferData()` |

> **Where this sits in the workshop.** This is the second half of batch
> correction. **Week 3** produced an integrated reduction on nine PBMC protocols
> and judged it by eye against labels fixed beforehand; this one produces a
> corrected expression matrix across two studies and judges it with a number. The
> companion Week 4 tutorials cluster this dataset and test genes between donors.
> Bulk deconvolution is **Week 5**.

# Setting up

```{r install}
# install.packages(c("Seurat", "RANN", "ggplot2", "patchwork"))
```

```{r libraries}
library(Seurat)
library(SeuratObject)
library(RANN)
library(ggplot2)
library(patchwork)
```

```{r clean_env}
rm(list = ls())
```

# The data: two studies, one brain region

**SEA-AD** (Seattle Alzheimer's Disease Brain Cell Atlas) profiled middle
temporal gyrus across an aged cohort spanning the full range of Alzheimer's
neuropathology. The file is the same one the other two Week 4 tutorials use: the
microglia and perivascular-macrophage subset, 40,000 nuclei from 89 donors, with
the authors' own `Supertype` annotation.

**ROSMAP**, as analysed by Sun *et al.* (2023), profiled immune cells from six
brain regions across 425 participants of the Religious Orders Study and Rush
Memory and Aging Project, and assigned each microglial nucleus one of thirteen
**states**, `MG0` through `MG12`. The file here is the middle temporal cortex
subset: around ten thousand nuclei from 48 individuals.

> **The region was matched on purpose, and this is the most important design
> decision in the tutorial.** ROSMAP's largest region by far is prefrontal
> cortex — six times as many microglia as middle temporal cortex, and the obvious
> thing to reach for. Pairing SEA-AD's middle temporal gyrus with ROSMAP's
> prefrontal cortex would have made **study** and **brain region** the same
> variable. Every difference between the two would then have two explanations,
> the integration would remove both, and no diagnostic could tell you which one
> you had lost. Week 3 made the same check with one line, `table(Method,
> Experiment)`, and found partial confounding in a dataset designed to be clean.
> Here the check is made before the data is chosen rather than after.
>
> Matching the region does not make the two studies comparable. It removes one
> named reason they are not.

```{r data_dir}
data_dir <- tempdir(check = TRUE)

# If you have already downloaded either file, point data_dir at the folder
# holding them and the downloads below will be skipped, e.g.:
# data_dir <- "/path/to/your/data"
```

```{r download_seaad}
options(timeout = 3600)

seaad_url <- paste0("https://www.dropbox.com/scl/fi/cei02rwh3gx61uu9zzeql/",
                    "seaad_microglia.RData",
                    "?rlkey=04erfwn3w2lrnr80pefwfh8fu&dl=1")
seaad_file <- file.path(data_dir, "seaad_microglia.RData")
seaad_size <- 525672784

if (!file.exists(seaad_file) || file.size(seaad_file) != seaad_size) {
  download.file(seaad_url, destfile = seaad_file, mode = "wb")
}

file.size(seaad_file) == seaad_size
```

```{r download_rosmap}
## TODO(Kevin): run private/code/rosmap/prep_rosmap_microglia_claude.R, upload
## the .RData it writes, and replace the two placeholder values below with the
## dl=1 share link and the byte count the script prints. This chunk cannot run
## until then. Delete this comment afterwards.
rosmap_url <- paste0("https://www.dropbox.com/scl/fi/siecgiljoks48pkin42ts/",
                     "rosmap_microglia.RData",
                     "?rlkey=vwnmoeynhbsuxkiumz36hy301&dl=1")
rosmap_file <- file.path(data_dir, "rosmap_microglia.RData")
rosmap_size <- 1507998516

if (!file.exists(rosmap_file) || file.size(rosmap_file) != rosmap_size) {
  download.file(rosmap_url, destfile = rosmap_file, mode = "wb")
}

file.size(rosmap_file) == rosmap_size
```

> **`options(timeout = 3600)`, `dl=1`, and `mode = "wb"`** are the same three
> download details the other Week 4 tutorials open with, and they matter for the
> same reasons: R's default 60-second timeout truncates a large file and reports
> it as a *warning*; a Dropbox link copied from the Share button ends in `dl=0`
> and serves an HTML preview page; and without binary mode the file is silently
> corrupted on Windows. The size comparison is there to check the bytes you got
> are the bytes you wanted.

```{r load_data}
load(seaad_file)
load(rosmap_file)

ls()
```

> **`load()` does not tell you what it is about to define.** It restores whatever
> names were in the file when it was saved, straight into your environment, and
> silently overwrites anything already using those names. The SEA-AD file defines
> `seurat_obj`; the ROSMAP file defines `rosmap_obj`. Had both used `seurat_obj`,
> the second `load()` would have destroyed the first with no message at all.
> `ls()` after a `load()` is a one-second habit worth having. (`readRDS()` avoids
> the problem entirely by returning its object instead of assigning it, which is
> a good reason to prefer `.rds` over `.RData` for anything you distribute.)

```{r rename_objects}
seaad <- seurat_obj
rosmap <- rosmap_obj
rm(list = c("seurat_obj", "rosmap_obj")); gc(TRUE)

seaad
rosmap
```

```{r inspect_seaad}
SeuratObject::Layers(seaad[["RNA"]])
length(unique(seaad$donor_id))
table(seaad$Supertype)
```

```{r inspect_rosmap}
SeuratObject::Layers(rosmap[["RNA"]])
length(unique(rosmap$donor_id))
table(rosmap$state)
table(rosmap$sequencing_batch)
table(rosmap$ad_diagnosis)
```

Two asymmetries are already visible and neither goes away.

SEA-AD arrives **log-normalized** (a `counts` layer and a `data` layer); ROSMAP
arrives as **counts only**. And the two studies do not use the same vocabulary
for cell identity: SEA-AD has eight `Supertype` labels built around
microglia/PVM subdivisions, ROSMAP has thirteen `state` labels built around
activation programs. Neither is wrong, and there is no dictionary between them.
That is the normal situation when you put two studies side by side, and the last
section of this tutorial is about what you can do with it.

# Making the two objects comparable

Three things have to be reconciled before any integration method will touch
them: barcodes, genes, and size.

## Barcodes

```{r rename_cells}
utils::head(Seurat::Cells(seaad), 3)
utils::head(Seurat::Cells(rosmap), 3)

seaad <- Seurat::RenameCells(seaad, add.cell.id = "seaad")
rosmap <- Seurat::RenameCells(rosmap, add.cell.id = "rosmap")

utils::head(Seurat::Cells(seaad), 3)
utils::head(Seurat::Cells(rosmap), 3)
```

Cell barcodes are drawn from a fixed 10x whitelist, so two experiments will share
thousands of them. `merge()` would deduplicate silently by appending suffixes,
and you would end up with an object whose cell names no longer match the ones in
any table you saved earlier.

```{r label_dataset}
seaad$dataset <- "SEA-AD"
rosmap$dataset <- "ROSMAP"
```

> **Record the batch as metadata, not as a prefix on the barcode.** It is
> tempting to recover it later with something like
> `strsplit(cell_name, "_")[[1]][1]`, and that works right up until a barcode
> contains an underscore of its own — at which point the parse quietly returns
> the wrong answer for a subset of your cells. One extra column costs nothing.

## Genes

```{r common_genes}
length(rownames(seaad))
length(rownames(rosmap))

common_genes <- intersect(rownames(seaad), rownames(rosmap))
length(common_genes)

length(setdiff(rownames(seaad), common_genes))
length(setdiff(rownames(rosmap), common_genes))
```

```{r subset_genes}
seaad <- subset(seaad, features = common_genes)
rosmap <- subset(rosmap, features = common_genes)

dim(seaad)
dim(rosmap)
```

> **An intersection is a lossy join, and the loss is not symmetric.** The two
> studies aligned to different genome annotation releases, so a gene can be
> missing from one side because it was renamed, because it was retired, or
> because that study filtered it out. Look at the two `setdiff()` counts above:
> whichever side loses more is the side whose gene list was longer, which usually
> means it kept features the other one discarded — antisense transcripts,
> lncRNAs, and unnamed `ENSG` loci.
>
> Anchors are matched by gene *name*, so the row order after subsetting does not
> matter. What does matter is that a gene you care about can simply be absent
> from `common_genes`, and nothing will tell you. If a specific marker is central
> to your question, check for it by name before you spend an hour integrating.

## Size

```{r downsample_seaad}
ncol(seaad)
ncol(rosmap)

set.seed(10)
seaad_cells <- sample(Seurat::Cells(seaad), size = 10000)
seaad <- subset(seaad, cells = seaad_cells)

ncol(seaad)
ncol(rosmap)
```

> **Anchor-based integration is sensitive to gross imbalance in dataset size.**
> Anchors are mutual nearest neighbours; a cell in the small dataset has far more
> candidate partners in the large one than the other way round, so the
> correction is estimated much better in one direction than the other. Forty
> thousand against ten thousand is a four-to-one ratio, and downsampling is the
> cheapest way to remove that as an explanation for whatever we see.
>
> It also makes the mixing score below interpretable: with two batches of equal
> size, "perfectly mixed" means half of every cell's neighbours come from the
> other study, and 0.5 is a number you can hold in your head. And it makes the
> tutorial knit in a workshop-sized amount of time — which is a real reason, just
> not a methodological one, and the two should not be confused.
>
> On your own data, downsampling to balance is a defensible choice you should
> **state in your methods**, not a default. Throwing away 30,000 measured nuclei
> is not free.

## Normalization and variable genes, within each study

```{r normalize_each}
seaad <- Seurat::NormalizeData(seaad, verbose = FALSE)
rosmap <- Seurat::NormalizeData(rosmap, verbose = FALSE)

seaad <- Seurat::FindVariableFeatures(seaad, selection.method = "vst",
                                      nfeatures = 2000, verbose = FALSE)
rosmap <- Seurat::FindVariableFeatures(rosmap, selection.method = "vst",
                                       nfeatures = 2000, verbose = FALSE)

length(intersect(Seurat::VariableFeatures(seaad),
                 Seurat::VariableFeatures(rosmap)))
```

```{r integration_features}
integration_features <- Seurat::SelectIntegrationFeatures(
  object.list = list(seaad, rosmap),
  nfeatures = 2000,
  verbose = FALSE)

length(integration_features)
utils::head(integration_features, 20)
```

Note how few of the two 2,000-gene lists agree. Each study's variable genes
partly describe that study's technical character, which is exactly why the gene
selection has to happen **inside** each batch and then be combined —
`SelectIntegrationFeatures()` ranks a gene by how often it is variable across the
datasets, so a gene that is only variable because of one study's chemistry loses
to a gene that is variable in both.

> **Run `FindVariableFeatures()` on the naively merged object and you select the
> batch effect.** A gene that is high in study A and low in study B has enormous
> variance in the pooled data, which is precisely what "highly variable" is
> measuring. You then hand those genes to PCA and wonder why PC1 is the study.
> Seurat v5 makes the right thing the default — `merge()` keeps one layer per
> dataset and `FindVariableFeatures()` operates per layer — but the trap is still
> live in any code that joins the layers first.

# What the two studies look like uncorrected

```{r merge_objects}
merged <- merge(seaad, rosmap)
merged

SeuratObject::Layers(merged[["RNA"]])
table(merged$dataset)
```

The layers stayed separate. That is Seurat v5's representation of "these cells
came from different places", and it is what lets the pipeline below behave
per-batch without being told to.

```{r uncorrected_pipeline}
merged <- Seurat::NormalizeData(merged, verbose = FALSE)
merged <- Seurat::FindVariableFeatures(merged, selection.method = "vst",
                                       nfeatures = 2000, verbose = FALSE)
merged <- Seurat::ScaleData(merged, verbose = FALSE)

set.seed(10)
merged <- Seurat::RunPCA(merged,
                         features = Seurat::VariableFeatures(merged),
                         npcs = 30,
                         reduction.name = "pca.uncorrected",
                         verbose = FALSE)

set.seed(10)
merged <- Seurat::RunUMAP(merged,
                          reduction = "pca.uncorrected",
                          dims = 1:30,
                          reduction.name = "umap.uncorrected",
                          seed.use = 10,
                          verbose = FALSE)

Seurat::Reductions(merged)
```

Every reduction here is given an explicit `reduction.name`. The defaults are
`pca` and `umap`, and a second `RunPCA()` later in the document would overwrite
the first without comment — leaving you comparing a corrected embedding against
itself. Naming them costs one argument and removes the possibility.

```{r dimplot_uncorrected, fig.width=7, fig.height=5, cache=FALSE}
gg <- Seurat::DimPlot(merged, reduction = "umap.uncorrected",
                      group.by = "dataset", raster = TRUE)
gg <- gg + ggplot2::ggtitle("Uncorrected, coloured by study")
plot(gg)
```

Overplotting can hide a lot in a two-colour UMAP — whichever study is drawn
second sits on top. `split.by` draws one panel per batch on shared axes, which is
the honest version of the same figure.

```{r dimplot_uncorrected_split, fig.width=9, fig.height=4.5, cache=FALSE}
gg <- Seurat::DimPlot(merged, reduction = "umap.uncorrected",
                      group.by = "dataset", split.by = "dataset",
                      raster = TRUE)
plot(gg)
```

## Measuring the separation instead of describing it

Week 3 judged its integration by eye and said so. A picture is genuinely
informative here, but it is also the thing most easily talked into agreeing with
you, and "the batches look reasonably mixed" is not a sentence a reviewer can
check. The number below is cheap and it is the core idea behind the published
integration metrics.

For each cell, find its `k` nearest neighbours in the embedding and ask what
fraction of them came from the *other* study.

```{r mixing_function}
neighbor_mixing <- function(embedding_mat, batch_vec, k = 30){
  # nn2 returns each point as its own first neighbour, so drop column 1.
  nn_out <- RANN::nn2(data = embedding_mat, k = k + 1)
  nn_idx <- nn_out$nn.idx[, -1, drop = FALSE]

  # batch_vec[nn_idx] unwinds column-major, so rebuild the matrix with the same
  # nrow to keep cell i on row i.
  neighbor_batch_mat <- matrix(batch_vec[nn_idx], nrow = nrow(nn_idx))

  rowMeans(neighbor_batch_mat != batch_vec)
}
```

```{r mixing_uncorrected}
dataset_vec <- merged$dataset
dataset_share <- table(dataset_vec) / length(dataset_vec)
dataset_share

# Under perfect mixing, a cell's neighbours are a random sample of everything
# else, so the expected other-study fraction is one minus its own study's share.
1 - dataset_share

mixing_uncorrected <- neighbor_mixing(
  embedding_mat = Seurat::Embeddings(merged, "pca.uncorrected")[, 1:30],
  batch_vec = dataset_vec,
  k = 30)

tapply(mixing_uncorrected, dataset_vec, stats::median)
```

Compare the two: the median observed against the value the second line predicts.
A score near zero means a cell's thirty nearest neighbours in a 30-dimensional
space all came from its own study — that the strongest thing about a nucleus in
this embedding is which lab sequenced it.

> **A mixing score alone rewards vandalism.** A method that mapped every cell to
> the same point would land exactly on the null for both studies — a perfect
> score, from an embedding containing nothing at all. Mixing is only half of
> the criterion; the other half is that biological structure survived, which is
> why the published metrics come in pairs — kBET or LISI for batch mixing,
> silhouette or cluster purity on known labels for preserved signal — and why
> Week 3 insisted on holding a cell-type label fixed *before* integrating.
> Read this number only alongside a picture that still has structure in it.

# Integrating so that the expression matrix is corrected too

Week 3's `IntegrateLayers()` takes an object and returns a reduction. The older
two-step route takes a **list** of objects and returns a new object carrying a
second assay.

```{r find_anchors}
integration_anchors <- Seurat::FindIntegrationAnchors(
  object.list = list(seaad, rosmap),
  anchor.features = integration_features,
  reduction = "cca",
  dims = 1:30,
  verbose = FALSE)

class(integration_anchors)
```

This is the slow step — expect several minutes. Canonical correlation analysis
finds directions of variation *shared* between the two datasets rather than
maximal within either; pairs of cells that are mutual nearest neighbours in that
shared space become **anchors**. An anchor is the method's assertion that *this*
SEA-AD nucleus and *that* ROSMAP nucleus are the same kind of cell, and
everything downstream inherits the quality of those assertions.

```{r integrate_data}
integrated <- Seurat::IntegrateData(anchorset = integration_anchors,
                                    new.assay.name = "integrated",
                                    dims = 1:30,
                                    verbose = FALSE)

integrated
Seurat::Assays(integrated)
Seurat::DefaultAssay(integrated)
```

For each anchor pair, Seurat computes a correction vector — the difference
between the two cells' expression profiles — and then applies a
distance-weighted average of nearby anchors' correction vectors to every cell in
the query dataset. The result is a matrix of the same shape as expression data,
made of expression data, that is not expression data.

> **`IntegrateData()` changes `DefaultAssay()` and does not say so.** After that
> one line, every function that does not name an assay explicitly — `FindMarkers()`,
> `FeaturePlot()`, `AverageExpression()`, `FindVariableFeatures()`, all of them —
> is reading corrected values instead of your measurements. This is the single
> most common way the mistake this tutorial is about actually happens: nobody
> decides to test genes on integrated data, they just never set `DefaultAssay()`
> back.

```{r join_rna_layers}
SeuratObject::Layers(integrated[["RNA"]])

integrated[["RNA"]] <- SeuratObject::JoinLayers(integrated[["RNA"]])

SeuratObject::Layers(integrated[["RNA"]])
```

> **`JoinLayers()` on the whole object fails here, and the error does not explain
> itself.** Week 3 wrote `SeuratObject::JoinLayers(obj)`, which is correct when
> every assay is a v5 assay. `IntegrateData()` builds the `integrated` assay as an
> **old-style v3 `Assay`**, which has no layers to join and no `JoinLayers`
> method, so the object-level call stops with
> `no applicable method for 'JoinLayers' applied to an object of class`
> `"c('Assay', 'KeyMixin')"` — an error that names a class you never asked for
> and does not mention integration at all. Joining the RNA assay **by name**, as
> above, sidesteps it. An object holding one v5 assay and one v3 assay is a
> normal state for Seurat 5 and worth being able to recognize.

# What is actually inside the `integrated` assay

```{r inspect_integrated_assay}
class(integrated[["integrated"]])
SeuratObject::Layers(integrated[["integrated"]])

dim(integrated[["RNA"]])
dim(integrated[["integrated"]])
```

Three facts, and each is a trap on its own.

**It is a v3 `Assay`, and its only layer is `data`.** There is no `counts` layer,
because there are no corrected counts — the correction happens in
log-normalized space and the result cannot be un-logged into integers. Anything
downstream that reaches for counts, which includes every count-based test in
existence, cannot use this assay at all. It will not error; `GetAssayData()` will
warn about a missing layer and hand back `NULL`.

**Most of your genes are gone.** The integrated assay has one row per
*anchor feature*, so roughly two thousand genes instead of the full shared gene
list. A gene you did not select as variable does not exist here. You can force
extra genes in with `IntegrateData(features.to.integrate = ...)`, and it is worth
knowing that argument exists, but "correct a gene I already know I care about,
then look at it" is a workflow with an obvious circularity problem.

```{r integrated_values}
integrated_data <- SeuratObject::LayerData(integrated[["integrated"]],
                                           layer = "data")
rna_data <- SeuratObject::LayerData(integrated[["RNA"]], layer = "data")

stopifnot(all(colnames(integrated_data) == colnames(rna_data)))

range(rna_data)
range(integrated_data)
sum(integrated_data < 0) / length(integrated_data)
```

**And the values go negative.** This is the part to hold on to. Log-normalized
expression is `log1p()` of a non-negative quantity, so it is bounded below by
zero by construction; a negative number in that matrix is not a low expression
level, it is not an expression level at all. It is the residue of a subtraction
between two cells that a nearest-neighbour search decided were comparable.

```{r zero_filling}
marker_candidates <- c("P2RY12", "CX3CR1", "CSF1R", "C1QB", "APOE", "SPP1")
marker_gene <- marker_candidates[marker_candidates %in%
                                   rownames(integrated_data)][1]
marker_gene

sum(rna_data[marker_gene, ] == 0)
sum(integrated_data[marker_gene, ] == 0)
```

Note the second habit worth stealing from that chunk: the marker is *chosen from
what is present* rather than hard-coded. A gene that survived the intersection
can still have been dropped by the anchor-feature selection, and a hard-coded
name would fail three chunks later with `subscript out of bounds`.

The zero counts are the other half of the story. A gene that was undetected in a
cell — a structural zero, a dropout, a real absence, you cannot tell which — has
been given a non-zero value by the correction. Every zero-inflation model, every
detection-rate covariate, every `min.pct` filter is now reading a number that was
imputed rather than measured.

```{r scatter_rna_vs_integrated, fig.width=7, fig.height=5, cache=FALSE}
compare_df <- data.frame(
  rna = rna_data[marker_gene, ],
  integrated = integrated_data[marker_gene, ],
  dataset = integrated$dataset)

gg <- ggplot2::ggplot(compare_df,
                      ggplot2::aes(x = rna, y = integrated, color = dataset))
gg <- gg + ggplot2::geom_point(size = 0.3, alpha = 0.2)
gg <- gg + ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed")
gg <- gg + ggplot2::geom_hline(yintercept = 0, color = "red")
gg <- gg + ggplot2::labs(x = paste0(marker_gene, ", measured (log-normalized)"),
                         y = paste0(marker_gene, ", after integration"),
                         title = "The same gene, the same cells, two matrices")
plot(gg)
```

The dashed line is where the two would agree. The red line is zero: everything
below it is a value with no possible interpretation as an amount of RNA. Note
also that the vertical stripe at `x = 0` — the cells where this gene was not
detected — has been spread across a range of corrected values, and that the two
studies are shifted differently, which is the correction doing its job.

# Re-embedding on the corrected values

```{r embed_integrated}
Seurat::DefaultAssay(integrated) <- "integrated"

integrated <- Seurat::ScaleData(integrated, verbose = FALSE)

set.seed(10)
integrated <- Seurat::RunPCA(integrated,
                             npcs = 30,
                             reduction.name = "pca.integrated",
                             verbose = FALSE)

set.seed(10)
integrated <- Seurat::RunUMAP(integrated,
                              reduction = "pca.integrated",
                              dims = 1:30,
                              reduction.name = "umap.integrated",
                              seed.use = 10,
                              verbose = FALSE)

Seurat::Reductions(integrated)
```

```{r dimplot_comparison, fig.width=10, fig.height=4.5, cache=FALSE}
gg1 <- Seurat::DimPlot(merged, reduction = "umap.uncorrected",
                       group.by = "dataset", raster = TRUE) +
  ggplot2::ggtitle("Before")
gg2 <- Seurat::DimPlot(integrated, reduction = "umap.integrated",
                       group.by = "dataset", raster = TRUE) +
  ggplot2::ggtitle("After")
plot(gg1 + gg2)
```

```{r mixing_integrated}
mixing_integrated <- neighbor_mixing(
  embedding_mat = Seurat::Embeddings(integrated, "pca.integrated")[, 1:30],
  batch_vec = integrated$dataset,
  k = 30)

tapply(mixing_uncorrected, merged$dataset, stats::median)
tapply(mixing_integrated, integrated$dataset, stats::median)
1 - dataset_share
```

```{r mixing_histogram, fig.width=8, fig.height=4, cache=FALSE}
mixing_df <- rbind(
  data.frame(mixing = mixing_uncorrected, dataset = merged$dataset,
             stage = "before"),
  data.frame(mixing = mixing_integrated, dataset = integrated$dataset,
             stage = "after"))
mixing_df$stage <- factor(mixing_df$stage, levels = c("before", "after"))

gg <- ggplot2::ggplot(mixing_df,
                      ggplot2::aes(x = mixing, fill = dataset))
gg <- gg + ggplot2::geom_histogram(bins = 40, alpha = 0.6,
                                   position = "identity")
gg <- gg + ggplot2::facet_wrap(~ stage)
gg <- gg + ggplot2::geom_vline(xintercept = 0.5, linetype = "dashed")
gg <- gg + ggplot2::labs(x = "fraction of 30 neighbours from the other study",
                         y = "cells")
plot(gg)
```

The dashed line is the balanced-design null. Read the *shape*, not only the
median: a distribution piled at zero with a thin tail says most cells never left
their own study, and a broad distribution centred near the line says the mixing
is general rather than driven by a handful of cells. A bimodal one — a mixed
population and an unmixed one — usually means a cell type that only one study
captured, and that is worth finding before you interpret anything.

> **Residual separation is not automatically a failure.** Two studies with
> different dissections and different chemistries can genuinely capture different
> states, and "the correction was incomplete" and "one study saw something the
> other did not" look identical in a UMAP. Telling them apart needs marker genes,
> not more integration. Week 3 hit the same ambiguity and left it standing for the
> same reason.

# Going back to the counts

```{r back_to_rna}
Seurat::DefaultAssay(integrated) <- "RNA"
Seurat::DefaultAssay(integrated)

SeuratObject::Layers(integrated[["RNA"]])
```

This is the whole discipline in two lines. The corrected assay is still there,
the reductions built from it are still there, and everything that reports a
number about a gene now reads the measurements again.

The rule, stated once:

> **Cluster and visualize on the integrated output. Test on the original
> counts.**

The reason is not a convention and it is worth being able to say out loud.
Integration deliberately removed variation it judged to be batch, using anchors
that *assumed* the two studies contain the same cell types. A statistical test
takes its input to be measurements with a known noise model — Poisson-ish
counts, or something you are willing to treat as normal — and computes how
surprising the observed difference is under that model. The corrected values are
neither. They are reconstructions whose errors are correlated across cells in
exactly the pattern the anchors imposed, and which have already had between-study
differences shrunk toward zero. The p-value that comes out is a real number
computed from the wrong thing; it does not come with a warning, and it is usually
*smaller* than the honest one, because the correction removed the variance the
test was supposed to be measuring against.

The right way to handle batch in a test is to keep it in the model rather than
remove it from the data — a `~ batch + condition` design on pseudobulk, which is
the companion Week 4 tutorial's subject.

> **So what is the corrected matrix legitimately for?** It is a *presentation*
> and *input* layer, not an evidence layer.
>
> - Plotting a gene across batches — `FeaturePlot()`, `DotPlot()`, `VlnPlot()` —
>   where an uncorrected panel would mostly show you which study each cell came
>   from.
> - Heatmaps of a marker panel across studies, for the same reason.
> - Feeding a downstream method that wants a genes × cells matrix and has no
>   batch term of its own: some trajectory methods, some matrix-factorization and
>   gene-program methods.
>
> Everything on that list is descriptive. Nothing on it produces a p-value, a
> fold change, or a confidence interval. That is the line.

```{r featureplot_comparison, fig.width=10, fig.height=4.5, cache=FALSE}
Seurat::DefaultAssay(integrated) <- "RNA"
gg1 <- Seurat::FeaturePlot(integrated, features = marker_gene,
                           reduction = "umap.integrated", raster = TRUE) +
  ggplot2::ggtitle(paste0(marker_gene, ", measured"))

Seurat::DefaultAssay(integrated) <- "integrated"
gg2 <- Seurat::FeaturePlot(integrated, features = marker_gene,
                           reduction = "umap.integrated", raster = TRUE) +
  ggplot2::ggtitle(paste0(marker_gene, ", corrected"))

Seurat::DefaultAssay(integrated) <- "RNA"
plot(gg1 + gg2)
```

Two panels, one gene, one set of coordinates, one argument different — and the
argument was never written down anywhere in the figure. If you make plots from
both assays in the same analysis, put the assay in the title, every time.

# The alternative that never edits expression: label transfer

There is a second thing people usually want when they reach for integration, and
it does not require correcting anything. **What are my cells, in someone else's
vocabulary?** ROSMAP's nuclei have `state` labels; SEA-AD's have `Supertype`
labels; the question is how they correspond.

```{r transfer_anchors}
transfer_anchors <- Seurat::FindTransferAnchors(reference = seaad,
                                                query = rosmap,
                                                dims = 1:30,
                                                verbose = FALSE)

predictions_df <- Seurat::TransferData(anchorset = transfer_anchors,
                                       refdata = seaad$Supertype,
                                       dims = 1:30,
                                       verbose = FALSE)

dim(predictions_df)
utils::head(colnames(predictions_df))
utils::head(predictions_df[, c("predicted.id", "prediction.score.max")], 5)
```

```{r attach_predictions}
rosmap <- Seurat::AddMetaData(rosmap, metadata = predictions_df)

table(rosmap$predicted.id)
stats::quantile(rosmap$prediction.score.max)
```

Note what did **not** happen: `rosmap` still holds its own counts, unmodified.
The anchors were used to move a *label* rather than to move the data, so
everything downstream — pseudobulk aggregation, DESeq2, composition testing —
operates on measurements.

```{r label_crosstab}
label_tab <- table(rosmap$state, rosmap$predicted.id)
label_tab

round(label_tab / rowSums(label_tab), 2)
```

The row proportions are the readable version: for each ROSMAP state, where do
those nuclei land in SEA-AD's vocabulary? A row concentrated in one column is a
state the reference recognizes. A row spread across several is a state the
reference has no name for, and this is where the method's central limitation
shows.

> **Label transfer cannot say "I do not know".** Every query cell gets the
> reference's best-matching label, whether or not the reference contains anything
> like it. A microglial activation state that SEA-AD's supertypes do not resolve
> does not come back as unlabelled; it comes back as the nearest supertype, with
> a confident-looking integer count in a table. `prediction.score.max` is the only
> thing standing between you and that — cells whose best score is barely above
> their second-best are cells whose label you should not use. Look at the
> quantiles above, pick a floor, and say what it was.
>
> The same caution applies in reverse to the direction of the transfer. We made
> SEA-AD the reference because its annotation is the one being borrowed; running
> it the other way, with ROSMAP's thirteen states as the reference, is a different
> analysis and can disagree.

# What we glossed over

- **We used CCA and compared it with nothing.** `FindIntegrationAnchors()` takes
  `reduction = "rpca"` (faster, more conservative — the recommended choice when
  the datasets are large or share little), and `IntegrateLayers()` in **Week 3**
  reaches Harmony, FastMNN and scVI through one argument. Running two methods and
  comparing them with the mixing score above costs one line and an hour, and
  "we used CCA because it was in the tutorial" is not a methods section.
- **Real integration metrics.** The neighbour-mixing score here is the idea
  behind LISI, boiled down to something you can read in the source. The published
  benchmarks — kBET, iLISI/cLISI, silhouette by batch and by label, and the
  composite scores in Luecken *et al.* — are what you would report. All of them
  pair a mixing measure with a biology-preservation measure, for the reason given
  above.
- **`SCTransform`.** Both routes accept `normalization.method = "SCT"`, but the
  anchor route additionally needs `PrepSCTIntegration()` before
  `FindIntegrationAnchors()`, and skipping it produces a wrong answer rather than
  an error. The **Week 3** normalization tutorial covers what SCTransform changes.
- **More than two datasets.** With many batches, `FindIntegrationAnchors(reference
  = c(1, 2))` integrates everything against a designated subset instead of
  all-against-all, which is the difference between an afternoon and a week.
  `sample.tree` in `IntegrateData()` controls the merge order, which is not
  cosmetic.
- **`k.anchor`, `k.filter`, `k.weight`.** Left at defaults. `k.weight` is the one
  that will actually break on you: it must not exceed the number of cells in the
  smallest dataset, and a small batch produces an error deep inside
  `IntegrateData()` that does not name the argument.
- **How many dimensions.** Thirty, everywhere, taken from the vignette and never
  defended — the same unexamined choice Week 3 flagged. It sets how much
  structure both the anchor-finding and the UMAP can see.
- **Whether these two studies should be integrated at all.** They have no shared
  donors and no shared protocol, so "study" is confounded with everything about
  how the tissue was collected and processed. Integration will always produce
  output. Whether the aligned populations are the same biology is a claim the
  method cannot make for you.
- **Cleaning.** Both objects arrive filtered by their authors, so **Week 2**'s
  ambient RNA and doublet work is invisible here. It comes first on your own data,
  and it comes first *per study* — ambient RNA profiles are study-specific.

# References

- Stuart T, Butler A, Hoffman P, *et al.* (2019). "Comprehensive integration of
  single-cell data." *Cell* 177:1888–1902.
  [doi:10.1016/j.cell.2019.05.031](https://doi.org/10.1016/j.cell.2019.05.031)
  — the anchor method used throughout this tutorial.
- Hao Y, Stuart T, Kowalski MH, *et al.* (2024). "Dictionary learning for
  integrative, multimodal and scalable single-cell analysis" (Seurat v5).
  *Nature Biotechnology* 42:293–304.
  [doi:10.1038/s41587-023-01767-y](https://doi.org/10.1038/s41587-023-01767-y)
- Luecken MD, Büttner M, Chaichoompu K, *et al.* (2022). "Benchmarking
  atlas-level data integration in single-cell genomics." *Nature Methods*
  19:41–50.
  [doi:10.1038/s41592-021-01336-8](https://doi.org/10.1038/s41592-021-01336-8)
  — the source for kBET, LISI and the mixing-versus-preservation framing.
- Sun N, Victor MB, Park YP, *et al.* (2023). "Human microglial state dynamics
  in Alzheimer's disease progression." *Cell* 186:4386–4403.
  [doi:10.1016/j.cell.2023.08.037](https://doi.org/10.1016/j.cell.2023.08.037)
  — the ROSMAP microglia and the `MG0`–`MG12` states.
  Project page: [https://compbio.mit.edu/microglia_states/](https://compbio.mit.edu/microglia_states/)
- Gabitto MI, Travaglini KJ, Rachleff VM, *et al.* (2024). "Integrated multimodal
  cell atlas of Alzheimer's disease" (SEA-AD).
  [https://portal.brain-map.org/explore/seattle-alzheimers-disease](https://portal.brain-map.org/explore/seattle-alzheimers-disease)
- Prater KE, Lin KZ (2025). "All the single cells: single-cell
  transcriptomics/epigenomics experimental design and analysis considerations for
  glial biologists." *Glia* 73:451–473.
  [doi:10.1002/glia.24633](https://doi.org/10.1002/glia.24633)
  — §3.3 on batch correction covers this tutorial directly.
