---
title: "Week 4: Differential expression across donors, and what the genes mean"
author: "Kevin Z. Lin"
date: "2026-08-15"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE, autodep = TRUE)
knitr::dep_auto()
```

> **A note on `cache = TRUE`, and on the `autodep` line next to it.** This
> document caches every chunk, so the first knit is slow and later ones take a
> couple of minutes. Two things are worth knowing before you rely on that.
>
> First, the disk cost: `knitr` re-saves every object a chunk *modifies*, and
> single-cell objects are large, so `_cache/` reaches gigabytes. Every chunk here
> whose only output is a figure is marked `cache=FALSE`, because a `ggplot`
> object stores the environment it was built in and caching one can drag a copy
> of the whole Seurat object onto disk with it. Exclude `_cache/` from Dropbox,
> OneDrive or Google Drive syncing, and delete it if a knit is interrupted — a
> truncated cache reappears on the next run as `error reading from connection`.
>
> Second, and more dangerous: **by default `knitr` caches each chunk
> independently and does not re-run a chunk when the objects it depends on
> change.** Edit an upstream chunk, re-knit, and a downstream figure can silently
> remain the *old* figure — a document whose text and plots disagree, with
> nothing to warn you. `autodep = TRUE` plus `knitr::dep_auto()` makes `knitr`
> track which chunks read which objects and invalidate accordingly. It is not
> perfect. Before anything leaves your hands, delete `_cache/` and knit once from
> clean.

# Overview

This tutorial was made from the DESeq2 vignette,
[https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html),
the `clusterProfiler` book,
[https://yulab-smu.top/biomedical-knowledge-mining-book/](https://yulab-smu.top/biomedical-knowledge-mining-book/),
and the pseudobulk section of the Harvard Chan Bioinformatics Core single-cell
course. The diagnostic steps in between are the checks worth running before you
trust any donor-level result.

The companion Week 4 tutorial clustered microglia and asked *what distinguishes
one cluster from another*. This one asks a completely different question:

> **What is different between these people?**

Both questions are called "differential expression," both are answered with a
table of genes and p-values, and they are **not the same analysis**. Confusing
them is the single most common error in this corner of the field, and it is the
reason this session has two tutorials.

The difference is what counts as a replicate. When you ask what makes cluster 3
different from cluster 5, your replicates are cells, and you have thousands. When
you ask what makes a donor with Alzheimer's disease different from one without,
your replicates are **donors**, and you have dozens at best. Testing that second
question with cells as replicates treats 500 nuclei from one brain as 500
independent observations of that person's biology. They are not — they came from
one piece of tissue, processed in one library, from one individual. Do it anyway
and you get p-values of `1e-80` for genes that would not survive a look at the
donor-level averages.

The fix is **pseudobulk**: sum the counts across all cells of a given type within
each donor, producing one profile per donor, then analyze that with the bulk
RNA-seq tools that were built for exactly this design.

The plan:

| Stage | What it produces | Function |
|---|---|---|
| 1. Pick a population and check it is testable | A cells-per-donor table you looked at before deciding | `table()` |
| 2. Aggregate | One column per donor, in raw counts | `AggregateExpression()` |
| 3. Filter | A gene universe chosen on expression, not on variability | `rowSums()` |
| 4. Check the design | A full-rank model matrix, verified | `model.matrix()`, `qr()` |
| 5. Look before testing | Sample-sample correlation, PCA on the pseudobulk | `vst()`, `plotPCA()` |
| 6. Test | Log fold changes and p-values per gene | `DESeq()`, `results()` |
| 7. Shrink | Fold changes you can rank | `lfcShrink()` |
| 8. Ask what the genes do | Enriched biological processes | `gseGO()` |

> **Where this sits in the workshop.** Week 1 ran a whole pipeline on defaults
> and defended nothing. Week 2 cleaned the data. Week 3 normalized it, picked
> genes, and did batch correction part one. This session is the payoff: the
> companion tutorial clusters and labels, and this one tests. Bulk deconvolution
> is **Week 5**.

# Setting up

Everything here is on CRAN or Bioconductor. Install once, then leave these lines
commented.

```{r install}
# install.packages(c("Seurat", "ggplot2", "pheatmap", "RColorBrewer"))
# if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("DESeq2", "apeglm", "EnhancedVolcano",
#                        "clusterProfiler", "org.Hs.eg.db", "enrichplot",
#                        "variancePartition"))
```

```{r libraries}
library(Seurat)
library(SeuratObject)
library(DESeq2)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(variancePartition)
library(ggplot2)
library(pheatmap)
```

```{r clean_env}
rm(list = ls())
```

# The data

The dataset is a 40,000-nucleus microglia subset of **SEA-AD** (Seattle
Alzheimer's Disease Brain Cell Atlas), middle temporal gyrus, distributed as a
Seurat object. The companion clustering tutorial describes it in full; here we
need only three things from it — the raw counts, the donor each nucleus came
from, and the donor's neuropathology.

The file downloads itself into a temporary folder, so this tutorial runs as-is
with nothing to set up. It is **525 MB**, so expect the first run to take a few
minutes on a decent connection.

```{r data_dir}
data_dir <- tempdir(check = TRUE)

# If you have already downloaded seaad_microglia.RData, point data_dir at the
# folder holding it instead and the download below will be skipped, e.g.:
# data_dir <- "/path/to/your/data"
```

```{r download}
options(timeout = 3600)

data_url <- paste0("https://www.dropbox.com/scl/fi/cei02rwh3gx61uu9zzeql/",
                   "seaad_microglia.RData",
                   "?rlkey=04erfwn3w2lrnr80pefwfh8fu&dl=1")
data_file <- file.path(data_dir, "seaad_microglia.RData")
expected_size <- 525672784

if (!file.exists(data_file) || file.size(data_file) != expected_size) {
  download.file(data_url, destfile = data_file, mode = "wb")
}

file.size(data_file) == expected_size
```

> **Three details in that download, and every one of them has bitten someone.**
>
> **`options(timeout = 3600)`.** R's default download timeout is **60 seconds**,
> which is not enough for half a gigabyte on any normal connection. Without this
> line the transfer dies partway through with
> `downloaded length 106201070 != reported length 525672784` — and because that
> arrives as a *warning* rather than an error, a script can sail past it and fail
> later at `load()` with a message about a corrupt file. This is the single most
> common reason a large `download.file()` "doesn't work" in R.
>
> **`dl=1`, not `dl=0`.** Copying a Dropbox share link gives you `dl=0`, which
> serves the HTML preview page. You would download a web page, and `load()` would
> complain about a bad magic number.
>
> **`mode = "wb"`.** Forces a binary write. Without it the file is silently
> corrupted on Windows.
>
> The size comparison at the end is the point of all three: **verify the bytes you
> got are the bytes you wanted**, rather than trusting that no error appeared.

```{r load_data}
load(data_file)
seurat_obj
```

> **`load()` is not `readRDS()`.** `load()` drops whatever objects the file
> happened to contain into your workspace under whatever names they were saved
> with, and it does **not** return a value — `x <- load("f.RData")` assigns a
> character vector of names, not your data, which is a genuinely confusing first
> encounter. Run `ls()` after every `load()` to see what you actually got, and
> prefer `saveRDS()` / `readRDS()` in your own work: those handle exactly one
> object and let *you* name it.

```{r list_objects}
ls()
```

The metadata is where the clinical variables live. There are 43 columns; these
are the ones this tutorial uses.

```{r metadata_columns}
head(seurat_obj@meta.data[, c("donor_id", "Supertype", "assay", "sex",
                              "ADNC", "APOE4.status", "Cognitive.status")])
```

`ADNC` is the **Alzheimer's Disease Neuropathologic Change** score — the
autopsy-derived measure of how much AD pathology the brain actually had, on a
four-level scale plus a `Reference` category for the neurotypical donors used to
build the atlas.

```{r adnc_table}
table(seurat_obj$ADNC)
table(seurat_obj$assay)
```

## Two subsets, both deliberate

```{r subset_object}
seurat_obj <- subset(seurat_obj,
                     ADNC %in% c("High", "Intermediate", "Low", "Not AD"))
seurat_obj <- subset(seurat_obj, assay == "10x 3' v3")
seurat_obj
```

The first line drops the `Reference` donors. They are a different cohort recruited
under different criteria, and they have no ADNC score to compare — including them
would mean silently comparing across studies.

The second line keeps a single chemistry. SEA-AD ran both 10x 3' v3 and 10x
multiome, and **27 donors appear in both**. If you aggregate by donor without
splitting on assay, those donors get one profile that pools two different library
preparations; if you split on both, they contribute twice and are no longer
independent. Restricting to one chemistry avoids the question entirely, at the
cost of some cells. Week 3's integration material is about the other option.

```{r donor_count}
length(unique(seurat_obj$donor_id))
```

# Choosing a population to test

Pseudobulk is done **within a cell type**, not across all cells at once — the
whole point is to ask whether microglia of a given state differ between donors,
and averaging distinct states together would hide that. SEA-AD ships its own
fine-grained labels in `Supertype`.

Before picking one, look at how many cells each donor contributes to each. This
is the diagnostic that decides whether the analysis is possible at all, and it is
worth running before you have any attachment to a particular comparison.

```{r cells_per_supertype}
cells_tab <- table(seurat_obj$Supertype, seurat_obj$donor_id)
dim(cells_tab)

apply(cells_tab, 1,
      function(x) c(donors_with_10_or_more = sum(x >= 10),
                    median_cells = median(x),
                    max_cells = max(x)))
```

Read that table as a feasibility check, not a summary. `Micro-PVM_2` is present
in nearly every donor at a workable depth. `Micro-PVM_2_1-SEAAD` and
`Micro-PVM_4-SEAAD` are present in a handful of donors at a median of one cell
each — there is no donor-level comparison to be made there, and running DESeq2 on
them would produce a table of numbers that mean nothing.

> **This is why you look first.** A pipeline that loops DESeq2 over every cluster
> will happily return results for the clusters that have no business being
> tested, and those results look exactly like the real ones. The
> cells-per-donor-per-cluster table is the cheapest guard against that, and it
> takes one line.

We take `Micro-PVM_2`.

```{r subset_supertype}
seurat_obj <- subset(seurat_obj, Supertype == "Micro-PVM_2")
seurat_obj

summary(as.vector(table(seurat_obj$donor_id)))
```

# Aggregating to pseudobulk

`AggregateExpression()` sums raw counts within each group. Passing several
variables to `group.by` puts all of them in the resulting object's metadata,
which saves reassembling the donor table by hand.

```{r aggregate}
pseudo_seurat <- Seurat::AggregateExpression(
  seurat_obj,
  assays = "RNA",
  return.seurat = TRUE,
  group.by = c("donor_id", "APOE4.status", "ADNC", "sex"),
  verbose = FALSE)

pseudo_seurat
head(pseudo_seurat@meta.data)
```

One column per donor. Note what just happened to the size of the problem: tens of
thousands of cells became `r ncol(pseudo_seurat)` columns, and that number — not
the cell count — is the sample size every p-value below is based on.

> **Sum, not mean.** `AggregateExpression()` defaults to summing counts, and that
> is what you want. DESeq2 models raw counts with a negative binomial and derives
> its own size factors; handing it averages, or normalized values, breaks the
> assumption it uses to estimate dispersion. If you take one thing from this
> section: pseudobulk means *sum the counts*, then let the bulk tool normalize.

Pull out the count matrix and the metadata.

```{r extract_pseudobulk}
mat_pseudobulk <- SeuratObject::LayerData(pseudo_seurat,
                                          layer = "counts",
                                          assay = "RNA")
metadata_pseudobulk <- pseudo_seurat@meta.data

dim(mat_pseudobulk)
mat_pseudobulk[1:5, 1:3]
```

## How many cells is each column made of?

A pseudobulk column built from 800 nuclei and one built from 2 nuclei are not
comparable, and nothing downstream will tell you which is which. Record it.

```{r n_cells}
cells_per_donor <- table(seurat_obj$donor_id)
metadata_pseudobulk[, "n_cells"] <-
  as.vector(cells_per_donor[metadata_pseudobulk[, "donor_id"]])

summary(metadata_pseudobulk[, "n_cells"])
sort(metadata_pseudobulk[, "n_cells"])[1:8]
```

```{r filter_low_cells}
min_cells <- 10
keep_samples <- metadata_pseudobulk[, "n_cells"] >= min_cells
sum(!keep_samples)

mat_pseudobulk <- mat_pseudobulk[, keep_samples]
metadata_pseudobulk <- metadata_pseudobulk[keep_samples, ]
dim(mat_pseudobulk)
```

`min_cells = 10` is a judgment call, not a standard. The point is that some
threshold is needed and that yours should be stated.

## Preparing the variables

DESeq2 needs the variables in the design to be factors, and for each factor it
needs to know which level is the **reference** — the baseline that every other
level is reported against. If you do not say, R takes the alphabetically first
level.

```{r prepare_metadata}
metadata_pseudobulk[, "sex"] <- factor(metadata_pseudobulk[, "sex"])
metadata_pseudobulk[, "APOE4.status"] <-
  factor(metadata_pseudobulk[, "APOE4.status"])

adnc_vec <- rep("NonAD", nrow(metadata_pseudobulk))
adnc_vec[metadata_pseudobulk[, "ADNC"] %in% c("High", "Intermediate")] <- "AD"
metadata_pseudobulk[, "ADNC"] <- relevel(factor(adnc_vec), ref = "NonAD")

summary(metadata_pseudobulk[, c("sex", "APOE4.status", "ADNC")])
```

> **Why the reference level matters: it decides the direction, not the
> statistics.** For a two-level factor the fit is symmetric. Swap the reference
> and DESeq2 estimates the same dispersions, the same Wald statistic up to sign,
> and **identical p-values**. What changes is which way round the comparison is
> reported. Our two levels are `AD` and `NonAD`, and `AD` sorts first
> alphabetically — so without the `relevel()` above, DESeq2 would treat AD as the
> baseline and give you a coefficient named `ADNC_NonAD_vs_AD`. Every
> `log2FoldChange` in this tutorial would then carry the opposite sign: the
> right-hand side of the volcano plot would be genes higher in *controls*, the
> "upregulated in AD" list would be the downregulated one, and the GSEA at the
> end — which ranks genes by the signed Wald statistic — would report every
> pathway with its enrichment score flipped. Nothing errors, no number looks
> wrong, and the biology you write down is backwards. So the answer to "why does
> it matter if it is not the one I meant" is: it matters entirely for
> interpretation, and interpretation is the whole output.
>
> Two consequences. First, the fix is not to remember which level sorts first —
> it is to set the reference on purpose and then **read `resultsNames()`** after
> fitting, which spells the comparison out as `variable_level_vs_reference`. We
> do that below. Second, with more than two levels the reference does more than
> set a sign: it decides which contrasts exist as coefficients at all. With
> `Not AD` as the reference you get `High vs Not AD`, `Low vs Not AD` and so on,
> and `High vs Low` is not among them — you would have to request it explicitly
> with `results(dds, contrast = c("ADNC", "High", "Low"))`.

Collapsing four ADNC levels into two is a modelling decision with consequences:
it buys power and a fold change that is easy to describe, and it throws away the
ordering. `Low` and `Not AD` become the same thing; so do `High` and
`Intermediate`. An ordinal model, or ADNC as a numeric trend, is the alternative
and is not obviously worse.

## Check the alignment before you trust it

DESeq2 matches columns of the count matrix to rows of the metadata **by
position**, not by name. If they ever fall out of order, every result is wrong
and nothing errors.

```{r check_alignment}
all(colnames(mat_pseudobulk) == rownames(metadata_pseudobulk))
```

> **The check that has to be there.** Print a sample-order confirmation before
> every DE run — it costs one line, and this is why. A mis-joined
> metadata table produces a complete, plausible, entirely fictional result set.
> There is no downstream symptom.

# Choosing which genes to test

The tempting shortcut here is to reuse the 2,000 variable genes the clustering
workflow already selected. They are sitting right there in the object, and
testing 2,000 genes instead of 15,000 is faster and gives a gentler multiple-
testing correction. It is a trap, and it is worth naming.

> **Do not select genes on variability and then test them for differential
> expression.** `FindVariableFeatures()` ranks genes by how much they vary
> *across cells* — which includes the between-donor variation you are about to
> test. Choosing the top 2,000 on that criterion and then asking which of them
> differ between donor groups is selecting on the outcome. It also silently
> redefines the multiple-testing universe: your FDR is now computed over a set
> that was already enriched for what you are looking for.

Filter on **detection** instead, which is orthogonal to the contrast: keep genes
seen in a reasonable number of donors. The equivalent filter can also be applied
at the single-cell level before aggregating.

```{r filter_genes}
min_samples <- ceiling(0.25 * ncol(mat_pseudobulk))
min_samples

genes_detected <- Matrix::rowSums(mat_pseudobulk >= 5) >= min_samples
sum(genes_detected)

mat_pseudobulk <- mat_pseudobulk[genes_detected, ]
dim(mat_pseudobulk)
```

# The design

```{r design}
design_formula <- ~ sex + APOE4.status + ADNC
design_mat <- model.matrix(design_formula, metadata_pseudobulk)

colnames(design_mat)
dim(design_mat)
```

`ADNC` is the contrast of interest; `sex` and `APOE4.status` are in the model to
be adjusted for. Everything in the formula that is not the term you read out is
a covariate, and adding covariates costs degrees of freedom — with 80-odd donors
that is affordable, with 8 it is not.

A design matrix that is not full rank makes DESeq2 fail, or worse, silently
estimate something other than what you wrote. Check it.

```{r check_rank}
qr(design_mat)$rank
ncol(design_mat)
qr(design_mat)$rank == ncol(design_mat)
```

> **Do not "fix" this automatically.** The obvious repair is to drop design
> columns in a loop until the matrix inverts, and it is a repair you will find
> written into pipelines because it keeps them running on small unbalanced
> cohorts. The cost is that the model which was fitted is no longer the model
> that was specified, and unless the code says which columns it dropped, nothing
> in the output tells you that happened — you get a clean table of results for a
> question you did not ask. If
> your design is not full rank, the honest responses are to drop a covariate on
> purpose, collapse a level, or accept that the comparison is not identifiable in
> your data.

Full rank is a low bar, though. It says the coefficients are *identifiable*, not
that they are estimated from a balanced comparison. Cross-tabulate the covariates
against the contrast and look at the cells:

```{r confounding_check}
table(metadata_pseudobulk[, "ADNC"], metadata_pseudobulk[, "sex"])
table(metadata_pseudobulk[, "ADNC"], metadata_pseudobulk[, "APOE4.status"])
```

> **Read that second table before reading any APOE4 result.** Every APOE4-positive
> donor in this cohort is in the AD group; the non-AD group is entirely
> APOE4-negative. The design is still full rank — the AD group contains both
> genotypes, so the coefficient is estimable — but it is estimated **only from
> the within-AD contrast**. This is not a bug in the data; it is what APOE4 does,
> since carrying the allele is a major risk factor for having AD pathology at
> autopsy. It does mean that "adjusting for APOE4 status" here is doing something
> narrower than it sounds, and that a headline APOE4 effect from this design would
> need to be described very carefully. A design matrix check would not have caught
> it. A two-way table did.

# Looking at the data before testing it

```{r build_dds}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = mat_pseudobulk,
                                      colData = metadata_pseudobulk,
                                      design = design_formula)
dds
```

## Why the counts need transforming first

Every plot in this section is really a statement about **distance between
donors** — a correlation, a PCA. Distances are only meaningful if every gene gets
a comparable say in them, and on the raw count scale they do not. Count data is
heteroskedastic: a gene averaging 5,000 counts has a variance in the thousands, a
gene averaging 5 counts has a variance around 5. Correlate two donors on raw
counts and the answer is decided almost entirely by the few hundred loudest
genes.

The obvious fix, `log(x + 1)`, overcorrects at the other end. Down where a gene
has 1 count in one donor and 4 in another, that is pure sampling noise, but on
the log scale it is a two-fold change — as large a move as anything the highly
expressed genes make. Log-transformed data is dominated by its quietest, least
reliable genes.

A **variance-stabilizing transform** fits the actual mean–variance relationship
in *your* data and applies the transformation that flattens it, so that variance
is roughly constant across the whole expression range. High-count genes get
compressed, low-count genes get shrunk toward the gene's overall mean rather than
inflated. After it, a difference of 0.5 means about the same amount of evidence
whether the gene is loud or quiet, and a correlation between two donors is a
statement about their transcriptomes rather than about their sequencing depth.

`vst()` and `rlog()` are DESeq2's two implementations of that idea and are
interchangeable in what they are for. `vst()` fits a parametric curve to the
dispersion trend and is fast; `rlog()` shrinks each gene individually and handles
small experiments and wildly varying library sizes better, at a cost that grows
badly with sample size. DESeq2's own advice is `rlog()` below roughly 30 samples
and `vst()` above, and we have `r ncol(mat_pseudobulk)`.

```{r vst}
vsd <- DESeq2::vst(dds, blind = TRUE)
```

`blind = TRUE` matters: it tells the transform to ignore the design, so the
picture you are about to look at is not shaped by the hypothesis you are about to
test.

## What to look for in the correlation heatmap

```{r sample_correlation, fig.width=8, fig.height=7, cache=FALSE}
vsd_mat <- SummarizedExperiment::assay(vsd)
cor_mat <- cor(vsd_mat)

annotation_df <- data.frame(ADNC = metadata_pseudobulk[, "ADNC"],
                            sex = metadata_pseudobulk[, "sex"],
                            row.names = rownames(metadata_pseudobulk))

pheatmap::pheatmap(cor_mat,
                   annotation_col = annotation_df,
                   show_rownames = FALSE,
                   show_colnames = FALSE,
                   main = "Pseudobulk sample-sample correlation")
```

An 80 × 80 grid of colour is not something to judge by eye, so put numbers on the
same questions first.

```{r correlation_numbers, cache=FALSE}
cor_offdiag <- cor_mat[lower.tri(cor_mat)]
round(quantile(cor_offdiag, c(0, 0.05, 0.5, 0.95, 1)), 3)

mean_cor <- (rowSums(cor_mat) - 1) / (ncol(cor_mat) - 1)
round(sort(mean_cor)[1:5], 3)

metadata_pseudobulk[names(sort(mean_cor)[1:5]), "n_cells"]

cor(mean_cor, log10(metadata_pseudobulk[, "n_cells"]))
```

Now the plot. There are four things worth looking for in it, and the reason to
name them is that a heatmap like this gets printed far more often than it gets
read.

**One: the overall level.** Every column is the same cell type from the same
brain region, so these profiles ought to be very similar to one another, and they
are — the middle 90% of the off-diagonal correlations run from
`r round(quantile(cor_offdiag, 0.05), 2)` to
`r round(quantile(cor_offdiag, 0.95), 2)`. A version of this plot centred on 0.3
would not be telling you which donors differ; it would be telling you something
broke upstream. This is a *did anything go wrong* plot before it is anything
else.

**Two: individual samples that sit apart from everything.** These are what the
plot is best at. Look for a row and column that is uniformly cooler than the rest
of the grid, rather than for pairs — a donor that correlates poorly with the
whole cohort is a candidate to exclude; two donors that correlate poorly with
each other but fine with everyone else are just two donors. Three samples on the
left edge of this grid stand off from the rest, and `mean_cor` gives you their
names — around 0.83 average correlation against a cohort median of 0.92 — instead
of making you count squares. Finding them now matters because DESeq2 will not
mention them later: it estimates one dispersion per gene pooled across all
donors, so an aberrant donor widens that estimate and quietly costs power for
every gene in the experiment.

**Three: whether the block structure lines up with the annotation bars.** This is
the one people expect to be the payoff, and it usually is not. If AD and non-AD
pseudobulk profiles were fundamentally different, the `ADNC` bar would sort into
two blocks along the dendrogram. It does not — the green and red squares are
interleaved all the way across, and so are the sexes. That is the ordinary
outcome for a disease contrast in brain tissue, and it is worth absorbing before
reading a single p-value: whatever separates these donors most strongly, disease
status is not it. The variance decomposition at the end of this tutorial puts a
number on that; the heatmap gets you there for free in one line.

**Four: whether it lines up with something you did not want.** The structure that
*is* here — a dense, highly correlated core with paler samples trailing off it —
has to come from somewhere, and the usual suspects are technical: sequencing
depth, cells per donor, batch, PMI, RNA quality. Correlating each donor's average
correlation against its cell count is the cheap version of that check, and here
it comes back at **0.78**. That is the honest reading of this heatmap: it is
substantially a picture of how many nuclei each donor contributed. The three
samples flagged above are the three thinnest in the cohort, and a donor built
from few nuclei is *noisy*, not *strange* — which changes what you would do about
it, since excluding it is a decision about precision rather than about sample
quality.

> **Low correlation and outlying position are not the same thing.** PC1 of the
> same matrix, two plots down, correlates with that same cell count at only 0.19.
> There is no contradiction: sampling noise from a thin pseudobulk column is
> spread across all genes at once, which pulls that sample away from *everything*
> without pushing it in any particular direction. A correlation heatmap sees that
> and a PCA largely does not. It is a decent argument for looking at both, which
> is what the next two plots are for.

## The same questions, in a PCA

```{r pca_plot, fig.width=7, fig.height=5, cache=FALSE}
gg <- DESeq2::plotPCA(vsd, intgroup = "ADNC") +
  ggplot2::ggtitle("Pseudobulk PCA, coloured by ADNC")
plot(gg)
```

```{r pca_plot_ncells, fig.width=7, fig.height=5, cache=FALSE}
pca_df <- DESeq2::plotPCA(vsd, intgroup = "ADNC", returnData = TRUE)
pca_df[, "n_cells"] <- metadata_pseudobulk[, "n_cells"]

gg <- ggplot2::ggplot(pca_df, ggplot2::aes(x = PC1, y = PC2)) +
  ggplot2::geom_point(ggplot2::aes(colour = n_cells), size = 3) +
  ggplot2::scale_colour_viridis_c(trans = "log10") +
  ggplot2::ggtitle("The same PCA, coloured by cells per donor")
plot(gg)
```

Those two plots are the same points twice, and comparing them is the point: if
donors separate by how many cells they contributed rather than by their
pathology, the leading variation in your pseudobulk is a depth artifact and the
gene list will be about depth too.

Do not settle it by eye. Correlate the components against both.

```{r pca_correlations}
cor(pca_df[, "PC1"], log10(pca_df[, "n_cells"]))
cor(pca_df[, "PC2"], log10(pca_df[, "n_cells"]))

summary(aov(pca_df[, "PC1"] ~ metadata_pseudobulk[, "ADNC"]))
```

Here PC1 correlates with depth at 0.19 and separates the ADNC groups at
`p = 0.056` — so neither the artifact nor the biology dominates the leading
component. That is a mildly reassuring answer, and the reason to run the
check is that the alternative answers are the ones that would have changed what
you do next.

# Testing

```{r run_deseq}
dds <- DESeq2::DESeq(dds)
DESeq2::resultsNames(dds)
```

`resultsNames()` is worth reading rather than assuming. Each name is a coefficient
in the fitted model, and the one you want is spelled out in
`variable_level_vs_reference` form — evidence that `relevel()` did what you
intended.

```{r results}
deseq2_res <- DESeq2::results(dds, name = "ADNC_AD_vs_NonAD")
summary(deseq2_res)
```

```{r results_head}
deseq2_res_ordered <- deseq2_res[order(deseq2_res$padj), ]
head(as.data.frame(deseq2_res_ordered), 10)
```

## Shrinking the fold changes

A gene with few counts can post a huge fold change on almost no evidence: 3
counts in the non-AD donors against 30 in the AD donors is a ten-fold change, and
it is also nothing at all. Ranking by raw `log2FoldChange` puts genes like that
at the top of the table, above genes measured in the thousands that moved by a
solid and believable 40%.

`lfcShrink()` is the standard answer to that. It fits a prior to the distribution
of effect sizes across all genes and reports each gene's fold change under that
prior. A gene with thousands of counts has a tight likelihood and barely moves; a
gene with a handful has a wide one and gets pulled toward zero, because the data
never had much to say about it. `type = "apeglm"` is the current recommended
estimator; note it requires `coef =` naming a coefficient from
`resultsNames()` and will not take an arbitrary `contrast =`.

```{r shrink}
deseq2_shrunk <- DESeq2::lfcShrink(dds,
                                   coef = "ADNC_AD_vs_NonAD",
                                   type = "apeglm")
head(as.data.frame(deseq2_shrunk[order(deseq2_shrunk$padj), ]), 10)
```

```{r ma_plots, fig.width=9, fig.height=4.5, cache=FALSE}
par(mfrow = c(1, 2))
DESeq2::plotMA(deseq2_res, ylim = c(-3, 3), main = "Unshrunk")
DESeq2::plotMA(deseq2_shrunk, ylim = c(-3, 3), main = "Shrunk (apeglm)")
par(mfrow = c(1, 1))
```

The two panels are the same genes, and the difference between them is the whole
point. On the left, the low-count end of the axis is a wide fan: genes averaging
5 to 50 counts reaching a log2 fold change of ±2, none of them significant, all
of them ahead of the real hits if you sort by effect size. On the right that fan
has collapsed onto the zero line, while the blue significant genes — which sit at
means of a few hundred — have barely moved.

> **Yes, this is standard, and if you have not seen it, here is the likely
> reason.** Shrunken fold changes are what the DESeq2 vignette recommends for MA
> plots, volcano plots, and any ranking or thresholding of genes by effect size.
> They used to be automatic: until DESeq2 1.16, `results()` applied the shrinkage
> itself by default (`betaPrior = TRUE`), so a great deal of published work and
> a great many older tutorials report shrunken fold changes without ever
> mentioning it. That default was pulled out into a separate `lfcShrink()` call
> so that the *estimate* you report and the *test* you report stay visibly
> separate steps — which is clearer, at the cost that you now have to know the
> step exists. Reporting unshrunk fold changes is not an error, and plenty of
> papers do; it just means a six-fold change in your table may be one
> lightly-observed gene.

The p-values are identical between the two — shrinkage changes the effect size
estimate, not the test. Use the unshrunk table to decide *what is significant*,
and the shrunk one to decide *what is large*. That division is also why the GSEA
at the end of this tutorial ranks on the Wald statistic from the unshrunk
results: the statistic already carries the same information shrinkage adds, in
that it divides each fold change by its own standard error.

```{r volcano, fig.width=9, fig.height=7, cache=FALSE}
gg <- EnhancedVolcano::EnhancedVolcano(
  deseq2_shrunk,
  lab = rownames(deseq2_shrunk),
  x = "log2FoldChange",
  y = "pvalue",
  pCutoff = 0.05,
  FCcutoff = 0.25,
  title = "Micro-PVM_2, AD vs non-AD",
  subtitle = "Pseudobulk by donor, adjusted for sex and APOE4 status")
plot(gg)
```

> **The volcano's y-axis is the raw p-value, and its cutoff line is drawn at a
> raw p-value too.** That is the package default and it is not the threshold you
> should report. Read significance off `padj`.

# Interrogating the answer

You now have a ranked gene list, and the temptation is to go straight from here
to a volcano plot to a slide. **The step that gets skipped is reading the gene
names.** It is easy to skip precisely because nothing prompts you to do it: the
table is sorted, the p-values are small, the volcano looks like a volcano, and
the output of `results()` gives no sign that anything is amiss. Sorting a column
is not the same as looking at what is in it. So before this becomes a figure,
read the top twenty genes and ask whether they have something in common that has
nothing to do with your hypothesis.

```{r top_genes}
top_genes <- rownames(deseq2_res_ordered)[1:20]
top_genes

sum(grepl("^MT-", top_genes))
```

Seven of the top ten genes, and eight of the top twenty, are mitochondrial. That
should stop you, because **these are nuclei**. A nucleus does not contain
mitochondria, so mitochondrial transcripts in a snRNA-seq count matrix are
ambient RNA — soup from lysed cells captured alongside the nucleus. Week 2 made
this point in the other direction — the ambient contamination fraction in nuclei
can be *estimated from* the mitochondrial genes, precisely because their true
expression is known to be zero.

So the strongest "AD versus non-AD" signal in this analysis is a candidate
artifact: if AD samples carry more ambient RNA — because of tissue quality, PMI,
dissociation, or anything else that differs systematically — then every
mitochondrial gene will come out higher in AD, with no biology involved.

Test it directly. The mitochondrial fraction is a per-donor quantity, so compute
it on the pseudobulk.

```{r mito_fraction}
mito_genes <- grep("^MT-", rownames(mat_pseudobulk), value = TRUE)
mito_genes

mito_fraction <- Matrix::colSums(mat_pseudobulk[mito_genes, ]) /
  Matrix::colSums(mat_pseudobulk)
metadata_pseudobulk[, "mito_fraction"] <- mito_fraction

summary(mito_fraction)
tapply(mito_fraction, metadata_pseudobulk[, "ADNC"], summary)
```

```{r mito_test}
wilcox.test(mito_fraction ~ metadata_pseudobulk[, "ADNC"])
```

```{r mito_plot, fig.width=6, fig.height=4.5, cache=FALSE}
plot_df <- data.frame(mito_fraction = mito_fraction,
                      ADNC = metadata_pseudobulk[, "ADNC"])

gg <- ggplot2::ggplot(plot_df,
                      ggplot2::aes(x = ADNC, y = mito_fraction)) +
  ggplot2::geom_boxplot(outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.15, alpha = 0.6) +
  ggplot2::labs(y = "Mitochondrial fraction of pseudobulk counts",
                title = "Ambient contamination is not balanced across groups") +
  ggplot2::theme_bw()
plot(gg)
```

The contamination is real and it is not balanced: the AD group's median
mitochondrial fraction is roughly double the non-AD group's (0.21% against
0.12%), and the difference survives a rank test at `p = 0.003`.

The plot adds something the medians do not. The two groups overlap almost
completely at the bottom; what separates them is a **tail of about a dozen AD
donors** running up to 1.4%, with no non-AD counterpart above 0.32%. So this is
not "AD samples are uniformly dirtier" — it is a subset of AD samples that are,
and they are enough to move a group mean. That distinction matters for what you
would do about it: a covariate treats it as a smooth gradient, whereas the
picture suggests a handful of samples you might simply exclude and re-run.

Put it in the model as a covariate and re-test.

```{r deseq_adjusted}
dds_adj <- DESeq2::DESeqDataSetFromMatrix(
  countData = mat_pseudobulk,
  colData = metadata_pseudobulk,
  design = ~ sex + APOE4.status + mito_fraction + ADNC)

dds_adj <- DESeq2::DESeq(dds_adj)
DESeq2::resultsNames(dds_adj)

deseq2_res_adj <- DESeq2::results(dds_adj, name = "ADNC_AD_vs_NonAD")
summary(deseq2_res_adj)
```

```{r compare_adjusted}
deseq2_res_adj_ordered <- deseq2_res_adj[order(deseq2_res_adj$padj), ]
head(as.data.frame(deseq2_res_adj_ordered), 10)

sum(deseq2_res$padj < 0.05, na.rm = TRUE)
sum(deseq2_res_adj$padj < 0.05, na.rm = TRUE)

sum(grepl("^MT-", rownames(deseq2_res_adj_ordered)[1:20]))
```

**The adjustment did not do what you might expect, and that is worth sitting
with.** Overall the result shrinks hard — 20 genes at `padj < 0.05` become 5, and
the count at `padj < 0.1` falls from 50 to 14 — so the covariate is absorbing a
lot. But the mitochondrial genes are still eight of the top twenty and still at
the very top; `MT-ND3` moves from `padj = 0.013` to `padj = 0.041` and stays
first. Adding the covariate weakened the evidence without reassigning the
ranking.

> **Why this particular adjustment cannot settle it.** `mito_fraction` is
> computed *from the mitochondrial genes themselves*. For every other gene it is
> a legitimate technical covariate; for `MT-ND3` it is very nearly the outcome,
> so regressing one on the other is close to circular and the coefficient is not
> interpretable as "the AD effect net of contamination." A cleaner design would
> use a contamination estimate derived independently — SoupX's `rho` from Week 2,
> or an ambient profile estimated from empty droplets — and that requires going
> back to raw CellRanger output this object no longer carries.

> **What to take from this, and what not to.** We have not shown the
> mitochondrial signal is artefactual. We have shown three things: the ambient
> fraction differs by group, the top of the gene list is entirely composed of
> genes that ambient contamination would move, and the data in hand **cannot
> separate** the two explanations. That is a real result and it belongs in the
> methods section. It is also completely invisible unless someone reads the gene
> names.
>
> The general habit: when a DE result is topped by a coherent block of genes that
> share a mechanism unrelated to your hypothesis — mitochondrial, ribosomal,
> heat-shock, sex-chromosome, haemoglobin — treat that as a diagnostic finding
> about your samples before treating it as a finding about your biology.

The genes that are *not* mitochondrial are the ones to carry forward. `FKBP5`,
`TLN2`, `PRKCA` and `TXNIP` survive in both versions of the model, and `FKBP5` in
particular is a well-studied stress-response gene with an existing AD literature —
which makes it a hypothesis, not a conclusion.

# Why this had to be pseudobulk

Everything above can be done the other way — test the cells directly, ignoring
which donor they came from. That is one line, so let us run it and compare.

```{r single_cell_de}
Seurat::Idents(seurat_obj) <- "ADNC"
adnc_binary <- rep("NonAD", ncol(seurat_obj))
adnc_binary[seurat_obj$ADNC %in% c("High", "Intermediate")] <- "AD"
seurat_obj$adnc_binary <- adnc_binary
Seurat::Idents(seurat_obj) <- "adnc_binary"

cell_level_res <- Seurat::FindMarkers(seurat_obj,
                                      ident.1 = "AD",
                                      ident.2 = "NonAD",
                                      test.use = "wilcox",
                                      logfc.threshold = 0,
                                      min.pct = 0.1)
head(cell_level_res, 5)
```

```{r compare_counts}
sum(cell_level_res$p_val_adj < 0.05, na.rm = TRUE)
sum(deseq2_res$padj < 0.05, na.rm = TRUE)

min(cell_level_res$p_val, na.rm = TRUE)
min(deseq2_res$pvalue, na.rm = TRUE)
```

**887 genes versus 20.** The smallest cell-level p-value is `1.9e-269`; the
smallest pseudobulk p-value is `2.5e-08`. That is 261 orders of magnitude of
apparent evidence produced entirely by changing what you call a replicate — the
counts, the genes, and the donors are identical in the two analyses.

Note also that the cell-level top table is led by the same mitochondrial genes,
which is a useful reminder that pseudo-replication and confounding are separate
problems: fixing one does not fix the other.

```{r compare_scatter, fig.width=7, fig.height=5.5, cache=FALSE}
shared_genes <- intersect(rownames(cell_level_res), rownames(deseq2_res))
length(shared_genes)

compare_df <- data.frame(
  gene = shared_genes,
  cell_level = -log10(cell_level_res[shared_genes, "p_val"] + 1e-300),
  pseudobulk = -log10(deseq2_res[shared_genes, "pvalue"]))

gg <- ggplot2::ggplot(compare_df,
                      ggplot2::aes(x = pseudobulk, y = cell_level)) +
  ggplot2::geom_point(alpha = 0.2, size = 0.6) +
  ggplot2::geom_abline(slope = 1, intercept = 0, colour = "red") +
  ggplot2::labs(x = "-log10 p, pseudobulk (donors as replicates)",
                y = "-log10 p, cell-level Wilcoxon (cells as replicates)",
                title = "The same genes, the same data, two definitions of n") +
  ggplot2::theme_bw()
plot(gg)
```

The red line is `y = x`, where the two tests would agree. Essentially every gene
sits above it, and the line is pressed flat against the bottom of the panel
because the two axes differ by roughly two orders of magnitude: the pseudobulk
`-log10 p` tops out near 8, the cell-level one runs past 250. That whole vertical
gap is pseudo-replication.

> **This is not an argument that cell-level tests are always wrong.** They are
> the right tool for the question in the companion tutorial — what distinguishes
> one cluster of cells from another — where cells genuinely are the units. The
> error is using them for a question about donors. Mixed models that keep cells
> as observations while modelling donor as a random effect are a third option and
> a real area of method development.

# How much of the variation is what you care about?

Before reading a gene list, it is worth asking how much of the expression
variation in this dataset any of these variables explains. `variancePartition`
answers that per gene.

```{r variance_partition}
varpart_formula <- ~ ADNC + sex + APOE4.status + mito_fraction + n_cells

vsd_mat_top <- vsd_mat[order(apply(vsd_mat, 1, var),
                             decreasing = TRUE)[1:2000], ]

varpart_res <- variancePartition::fitExtractVarPartModel(
  vsd_mat_top, varpart_formula, metadata_pseudobulk)

head(variancePartition::sortCols(varpart_res))
```

> **Fixed effects or random effects is not a formatting choice here.** The
> formula above declares every variable as a fixed effect, and that decision is
> load-bearing. Written instead as `~ (1 | sex) + (1 | ADNC) + ...`, every
> categorical variable becomes a random effect — and a random effect estimated
> from **two** levels has almost no information about its own variance, so the
> fitted variance component collapses to zero. Run this analysis that way and
> `XIST`, a gene expressed
> almost exclusively in females, comes back with sex explaining ~0% of its
> variance and residual ~100%. The result is not flagged; it just looks like a
> null. `variancePartition`'s own guidance is to model categorical variables with
> few levels as **fixed** effects and reserve random effects for variables with
> many levels — donor in a repeated-measures design, batch across dozens of runs.

The sanity check is right there in the output: the most variable genes in this
pseudobulk are sex-chromosome and X-inactivation genes, so `sex` had better
account for nearly all of their variance.

```{r variance_partition_check}
sex_genes <- intersect(c("XIST", "UTY", "USP9Y", "RPS4Y1", "DDX3Y", "KDM5D"),
                       rownames(varpart_res))
varpart_res[sex_genes, ]
```

```{r variance_partition_plot, fig.width=7, fig.height=5, cache=FALSE}
gg <- variancePartition::plotVarPart(variancePartition::sortCols(varpart_res)) +
  ggplot2::ggtitle("Variance explained per gene, 2000 most variable")
plot(gg)
```

```{r variance_partition_summary}
round(100 * colMeans(varpart_res), 2)
round(100 * apply(varpart_res, 2, median), 2)
```

Two things to read off this.

**`Residuals` is the tallest box**, at 81% on average. Most of the gene-to-gene
variation between donors is not explained by any label we have, which is the
honest headline and a useful calibration before reading any gene list.

**And the ordering of the rest is the finding.** Averaged over the 2,000 most
variable genes: `mito_fraction` **11.0%**, `n_cells` 3.6%, `sex` 2.0%, `ADNC`
**1.2%**, `APOE4.status` 1.0%. The ambient-contamination covariate explains
roughly **nine times** as much variance as the disease variable this whole
analysis is about. The tall spike in the `sex` column is the sanity check
working — a handful of sex-chromosome genes at nearly 100%, and almost nothing
else.

That is the same conclusion the mitochondrial section reached, arrived at
independently and without reference to any p-value. It does not mean there is no
AD signal; it means the AD signal is small relative to a technical one, which is
exactly the situation where analysis choices decide the answer.

# What do the genes do?

A ranked gene list is not a finding. Gene set enrichment analysis asks whether
genes belonging to a known biological process sit systematically high or low in
your ranking.

The ranking statistic is a choice. Ranking by shrunk log fold change asks "what
changed most"; ranking by the Wald statistic asks "what changed most relative to
its uncertainty", which folds in how well each gene was measured.

```{r gsea_ranking}
teststat_vec <- deseq2_res[, "stat"]
names(teststat_vec) <- rownames(deseq2_res)
teststat_vec <- teststat_vec[!is.na(teststat_vec)]
teststat_vec <- sort(teststat_vec, decreasing = TRUE)

length(teststat_vec)
head(teststat_vec)
tail(teststat_vec)
```

```{r unmapped_ids}
sum(grepl("^ENSG", names(teststat_vec)))
```

> **Not every row name is a gene symbol.** This matrix mixes HGNC symbols with
> bare Ensembl IDs for features that had no symbol at annotation time — and two
> of them sit in the top five of the ranking. Passing `keyType = "SYMBOL"` means
> `gseGO()` silently drops every one of them. That is the right behaviour, but
> "silently" is the operative word: nothing tells you how much of your ranked list
> was discarded before the enrichment was computed. Count them yourself.

> **GSEA is randomized, and its p-values move between runs unless you say
> otherwise.** `gseGO()` takes a `seed` argument that is `FALSE` by default, so
> the permutation p-values shift a little every time you run it. Passing
> `seed = TRUE` is enough to fix that. We check that claim rather than trusting
> it, two chunks down.

```{r gsea}
gse <- clusterProfiler::gseGO(teststat_vec,
                              ont = "BP",
                              keyType = "SYMBOL",
                              OrgDb = "org.Hs.eg.db",
                              minGSSize = 10,
                              maxGSSize = 500,
                              seed = TRUE,
                              verbose = FALSE)

gse_df <- as.data.frame(gse)
dim(gse_df)
head(gse_df[, c("Description", "setSize", "NES", "pvalue", "p.adjust")], 10)
```

```{r gsea_reproducible}
gse_again <- clusterProfiler::gseGO(teststat_vec,
                                    ont = "BP",
                                    keyType = "SYMBOL",
                                    OrgDb = "org.Hs.eg.db",
                                    minGSSize = 10,
                                    maxGSSize = 500,
                                    seed = TRUE,
                                    verbose = FALSE)

identical(as.data.frame(gse_again)$pvalue, gse_df$pvalue)
```

```{r gsea_dotplot, fig.width=9, fig.height=7, cache=FALSE}
gg <- enrichplot::dotplot(gse, showCategory = 15) +
  ggplot2::ggtitle("GO biological process, ranked by Wald statistic")
plot(gg)
```

A positive normalized enrichment score means the set sits toward the AD end of the
ranking; negative means the non-AD end.

Read the top ten and notice what is in it. `regulation of hemopoiesis` and
`regulation of leukocyte differentiation` come out positive — those are myeloid
processes and microglia are myeloid cells, so that is coherent. But
`modulation of chemical synaptic transmission`, `regulation of trans-synaptic
signaling`, `axon development`, `axonogenesis` and
`cell morphogenesis involved in neuron differentiation` are all strongly
**negative**, and this dataset contains nothing but microglia.

> **Neuronal gene sets in a purely microglial analysis are a contamination
> readout, not a discovery.** Microglia sit in a dense neuropil; ambient RNA in
> cortical snRNA-seq is dominated by neuronal transcripts. Their coordinated
> movement across the ranking tells you how much neuronal soup each group's
> libraries carried. It is the same phenomenon as the mitochondrial block above,
> arriving through a different door — and GSEA is *especially* prone to surfacing
> it, because ambient RNA moves whole functional modules together, which is
> exactly the pattern GSEA is built to detect.

```{r gsea_split, fig.width=10, fig.height=8, cache=FALSE}
gg <- enrichplot::dotplot(gse, showCategory = 10, split = ".sign") +
  ggplot2::facet_grid(. ~ .sign)
plot(gg)
```

> **GSEA is testing thousands of overlapping gene sets.** GO terms are nested —
> a parent term contains its children — so a single real signal produces a column
> of related, individually "significant" rows. Read the top of the table as one
> finding with several names, not as several findings.

# What we glossed over

- **We never decontaminated.** This object ships as counts; no SoupX correction
  was applied, which is why the ambient signal is so visible in the results. The
  **Week 2** tutorials cover the correction, and the honest workflow runs it
  before any of this. Doing it here would need the raw droplet matrices, which
  this object does not carry.
- **Which cell state to test.** We took `Micro-PVM_2` because it was the only
  supertype present in nearly every donor. The companion Week 4 tutorial derives
  clusters from scratch; testing those instead makes the DE result depend on a
  clustering resolution, and that dependence is real and rarely reported.
- **Cells as replicates *and* donors as a random effect.** Mixed models
  (`glmmTMB`, `NEBULA`, `muscat`) sit between the two extremes shown here. They
  are the current research frontier and not settled enough to teach as a default.
- **Ordinal pathology.** We binarized ADNC. Braak stage, Thal phase, CERAD score
  and SEA-AD's `Continuous.Pseudo.progression.Score` are all in the metadata and
  all ordered.
- **Compositional change.** If AD brains have proportionally fewer homeostatic
  microglia, that is a finding — and it is invisible to this analysis, which
  tests within a fixed cell type. Cell-type proportions are their own question,
  touched on in the companion tutorial.
- **Batch correction and DE.** We avoided the question by taking one chemistry.
  Testing genes on integrated data is genuinely delicate: integration alters a
  reduction, not counts, so DE should return to raw counts — which is exactly
  what pseudobulk does. Batch belongs in the design formula, not in the counts.
- **Multiple testing across clusters.** Run this for eight supertypes and you
  have done eight families of tests. Nothing here adjusts for that.
- **Over-representation analysis versus GSEA.** We ran GSEA on the full ranking.
  Taking the significant genes and running a Fisher test against GO
  (`enrichGO()`) is the other standard approach; it needs a threshold and it
  discards the ranking.

# References

- Love MI, Huber W, Anders S (2014). "Moderated estimation of fold change and
  dispersion for RNA-seq data with DESeq2." *Genome Biology* 15:550.
  [doi:10.1186/s13059-014-0550-8](https://doi.org/10.1186/s13059-014-0550-8)
- Zhu A, Ibrahim JG, Love MI (2019). "Heavy-tailed prior distributions for
  sequence count data" (`apeglm`). *Bioinformatics* 35:2084–2092.
  [doi:10.1093/bioinformatics/bty895](https://doi.org/10.1093/bioinformatics/bty895)
- Squair JW *et al.* (2021). "Confronting false discoveries in single-cell
  differential expression." *Nature Communications* 12:5692.
  [doi:10.1038/s41467-021-25960-2](https://doi.org/10.1038/s41467-021-25960-2)
  — the paper behind the pseudo-replication section.
- Hoffman GE, Schadt EE (2016). "variancePartition." *BMC Bioinformatics* 17:483.
  [doi:10.1186/s12859-016-1323-z](https://doi.org/10.1186/s12859-016-1323-z)
- Subramanian A *et al.* (2005). "Gene set enrichment analysis." *PNAS*
  102:15545–15550.
  [doi:10.1073/pnas.0506580102](https://doi.org/10.1073/pnas.0506580102)
- Prater KE, Lin KZ (2025). "All the single cells: single-cell
  transcriptomics/epigenomics experimental design and analysis considerations for
  glial biologists." *Glia* 73:451–473.
  [doi:10.1002/glia.24633](https://doi.org/10.1002/glia.24633)
- Gabitto MI *et al.* (2024). "Integrated multimodal cell atlas of Alzheimer's
  disease" (SEA-AD). *Nature Neuroscience*.
  [https://portal.brain-map.org/explore/seattle-alzheimers-disease](https://portal.brain-map.org/explore/seattle-alzheimers-disease)
