---
title: "Week 2: Removing ambient RNA with SoupX (the manual route)"
author: "Kevin Z. Lin"
date: "2026-08-03"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This overview was made from [https://rawcdn.githack.com/constantAmateur/SoupX/204b602418df12e9fdb4b68775a8b486c6504fe4/inst/doc/pbmcTutorial.html](https://rawcdn.githack.com/constantAmateur/SoupX/204b602418df12e9fdb4b68775a8b486c6504fe4/inst/doc/pbmcTutorial.html).

This is the second SoupX tutorial. The first one ran the whole procedure on
autopilot: `load10X()` assembled everything, `autoEstCont()` produced a
contamination fraction, `adjustCounts()` cleaned the counts, and we trusted all
three. This document takes the same three steps apart and does the middle one by
hand.

There are two reasons to care, and both of them will apply to people in this
room.

**First, the autopilot depends on things your data may not have.** Everything
convenient in the first tutorial came from CellRanger's `analysis/` folder being
present and from `load10X()` finding it. If your experiment was not processed
with CellRanger — a Parse run, a plate-based protocol, or a count matrix a
collaborator emailed you — there is no `analysis/` folder, `load10X()` does not
apply, and `autoEstCont()` refuses to run without clusters. You need to know how
to assemble the object and attach the clustering yourself.

**Second, the automatic estimate can be wrong, and you should be able to check
it.** In the first tutorial the estimate came back at `rho = 0.01`, sitting
exactly on the floor of the estimator's own search range. That was a defensible
answer for a very clean dataset, but "the number landed on the boundary" is
precisely the situation where you want a second, independent estimate. The manual
route provides one.

We also switch to a **dirtier dataset** on purpose. The PBMC 1k data from the
first tutorial was about as clean as real data gets, which made the correction
almost invisible. Here we use the 10x **PBMC 4k** dataset, the same one the SoupX
authors use in their vignette. It comes out at **9.8% contamination** against the
first tutorial's 1%, and the difference is dramatic: there, the most affected gene
lost its counts in 14% of cells; here, ten genes lose theirs in half their cells
or more. You will be able to see the correction do something.

The plan:

| Stage | What we do by hand | Functions |
|---|---|---|
| Build the object | Read the two matrices ourselves instead of letting `load10X()` do it | `Read10X`, `SoupChannel`, `estimateSoup` |
| Attach the metadata | Read CellRanger's clusters and tSNE from CSV and set them explicitly | `setClusters`, `setDR` |
| Choose estimation genes | Decide which genes are safe to measure contamination with | `quickMarkers`, `plotMarkerDistribution` |
| Choose estimation cells | Decide which cells can be assumed not to express them | `estimateNonExpressingCells` |
| Estimate and correct | Compute `rho` from those choices, then clean | `calculateContaminationFraction`, `adjustCounts` |

This assumes you have been through the first tutorial. Concepts introduced there
— the soup, `rho`, the soup profile, why the correction is done cluster-wise —
are not re-explained here.

# Setting up

```{r libraries}
library(SoupX)
library(Matrix)
library(ggplot2)
```

```{r clean_env}
rm(list = ls())
```

# Getting the data

We use the 10x **PBMC 4k** dataset: about 4,300 peripheral blood mononuclear
cells from a healthy donor, processed with an older version of CellRanger.

We download three archives. Two are the count matrices; the third is
CellRanger's secondary analysis, which we will deliberately read by hand later.

```{r download}
tmp_dir <- tempdir(check = TRUE)
base_url <- "https://cf.10xgenomics.com/samples/cell-exp/2.1.0/pbmc4k/pbmc4k"

for (component in c("raw_gene_bc_matrices",
                    "filtered_gene_bc_matrices",
                    "analysis")) {
  destination <- file.path(tmp_dir, paste0(component, ".tar.gz"))
  download.file(paste0(base_url, "_", component, ".tar.gz"),
                destfile = destination)
  untar(destination, exdir = tmp_dir)
}
```

Note the folder names: `raw_gene_bc_matrices` and `filtered_gene_bc_matrices`,
not `raw_feature_bc_matrix` and `filtered_feature_bc_matrix` as in the first
tutorial. This dataset was produced by CellRanger 2.1, before 10x renamed these
outputs. Underneath each is a further folder named after the reference genome,
`GRCh38`. Small differences like this are the normal texture of working with
other people's data, and they are why hardcoding paths from a tutorial into your
own analysis tends to fail.

# Building the SoupChannel by hand

In the first tutorial a single call to `load10X()` did everything. Here we do it
in pieces, because seeing the pieces is the point.

`Read10X()` turns a folder of 10x files into a sparse count matrix. We call it
twice: once for the droplets that CellRanger called as cells (the **t**able **o**f
**c**ounts, `toc`), and once for every droplet including the empty ones (the
**t**able **o**f **d**roplets, `tod`).

```{r read_matrices}
toc <- Seurat::Read10X(file.path(tmp_dir, "filtered_gene_bc_matrices", "GRCh38"))
tod <- Seurat::Read10X(file.path(tmp_dir, "raw_gene_bc_matrices", "GRCh38"))

dim(toc)
dim(tod)
```

The two matrices have the same 33,694 genes but wildly different numbers of
columns: 4,340 cells in `toc` against **737,280** droplets in `tod`. That second
number is the full 10x barcode whitelist — every barcode the chemistry could
possibly have produced — so more than 99% of those droplets contain no cell at
all. That enormous empty majority is exactly what makes the soup measurable.

Now we assemble the object. `calcSoupProfile = FALSE` tells `SoupChannel()` not
to profile the soup on construction, and then we do it in a separate, visible
step. The two-step version is equivalent to the default, but it makes it obvious
that profiling the soup is a real computation and not something that comes free
with the object.

```{r build_channel}
sc <- SoupX::SoupChannel(tod, toc, calcSoupProfile = FALSE)
sc <- SoupX::estimateSoup(sc)
sc
```

`estimateSoup()` looks at droplets with fewer than 100 UMIs — its default
`soupRange = c(0, 100)` — and computes each gene's share of the counts in them.
That is step 1 of the method, now done explicitly.

```{r soup_profile}
head(sc$soupProfile[order(sc$soupProfile$est, decreasing = TRUE), ], n = 20)
```

`MALAT1` tops the list at about 3.3%, followed by `B2M`, `TMSB4X`, `EEF1A1` and
then a long run of ribosomal protein genes. The `counts` column shows how much
evidence sits behind each estimate — 88,000 `MALAT1` counts pooled across the
empty droplets, so these are not fragile numbers.

# Attaching clusters and dimension reduction after the fact

This section is the heart of the tutorial.

At this moment `sc` has counts and a soup profile, and **nothing else**. It has
no clusters, so `autoEstCont()` would fail on it and `adjustCounts()` would fall
back to a weaker cell-by-cell correction. We downloaded CellRanger's `analysis/`
folder, but we never pointed `load10X()` at it, so nothing was picked up
automatically. That is deliberate: we are going to attach the information
ourselves, exactly as you would if it had come from a Seurat run, a collaborator,
or any other outside source.

CellRanger writes its clustering as a two-column CSV.

```{r read_clusters}
clustering_df <- read.csv(file.path(tmp_dir, "analysis", "clustering",
                                    "graphclust", "clusters.csv"))
head(clustering_df)
table(clustering_df$Cluster)
```

`setClusters()` wants a vector of cluster labels **named by cell barcode**, not
a data frame. `setNames()` builds exactly that: it takes the values as its first
argument and the names as its second.

```{r set_clusters}
sc <- SoupX::setClusters(sc, setNames(clustering_df$Cluster,
                                      clustering_df$Barcode))
```

> **The mistake to avoid.** It is tempting to write
> `setNames(clustering_df$Cluster, rownames(clustering_df))`. After `read.csv()`,
> the row names of a data frame are just `"1"`, `"2"`, `"3"`, … — the barcodes
> live in the `Barcode` *column*, not in the row names. With row names, the
> labels have no barcode to match against; SoupX falls back to assigning them
> **by position**, which silently gives the right answer only if the CSV happens
> to be in the same order as the count matrix. It is here, but relying on that is
> how you get a plot that looks fine and is wrong. Name things by barcode and let
> the matching be explicit.

The same idea for the tSNE coordinates.

```{r read_tsne}
tsne_df <- read.csv(file.path(tmp_dir, "analysis", "tsne",
                              "2_components", "projection.csv"))
head(tsne_df)
```

Two small things to handle. The CSV's columns are named `TSNE-1` and `TSNE-2`,
but `read.csv()` converts hyphens to dots, so in R they arrive as `TSNE.1` and
`TSNE.2` — a good habit is to always `head()` a freshly read file rather than
assume its column names. And `setDR()` matches on row names, so we move the
barcodes there.

```{r set_dr}
rownames(tsne_df) <- tsne_df$Barcode
sc <- SoupX::setDR(sc, tsne_df[colnames(sc$toc), c("TSNE.1", "TSNE.2")])

head(sc$metaData)
```

Indexing with `tsne_df[colnames(sc$toc), ]` reorders the rows to match the count
matrix's columns. Even when the file is already in the right order — as it is
here — writing it this way means the code stays correct if it ever is not.

# A first look at the data

```{r plot_clusters, fig.width=7, fig.height=5}
dd <- sc$metaData[colnames(sc$toc), ]
dd$clusters <- as.factor(paste0("Cluster_", dd$clusters))

mids <- aggregate(cbind(TSNE.1, TSNE.2) ~ clusters, data = dd, FUN = mean)

gg <- ggplot2::ggplot(dd, ggplot2::aes(TSNE.1, TSNE.2)) +
  ggplot2::geom_point(ggplot2::aes(colour = clusters), size = 0.2) +
  ggplot2::geom_label(data = mids, ggplot2::aes(label = clusters)) +
  ggplot2::ggtitle("PBMC 4k, CellRanger graph-based clusters") +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 1)))
plot(gg)
```

Note that the coordinate columns are still called `TSNE.1` and `TSNE.2` here.
`setDR()` does not rename what you hand it — it copies your columns into
`sc$metaData` under their existing names and records which ones they are in
`sc$DR`. (The SoupX vignette calls them `RD1` and `RD2` only because its bundled
metadata happened to arrive with those names.) The plotting functions later in
this document read `sc$DR`, so they will find these coordinates whatever they are
called.

Note also that these clusters are **numbered, not named**. CellRanger does not
know what a B cell is. Deciding which cluster is which cell type is Week 4's
topic; for now the numbers are enough, because the contamination estimate needs
to know only which cells group together, not what they are.

# Choosing genes to estimate the contamination with

Here is the whole idea in one sentence: **if we can find a gene that a cell
definitely does not express, then every count of that gene in that cell must have
come from the soup**, and comparing that count to the soup profile gives us
`rho` for that cell.

So we need genes that are confidently *off* in some identifiable set of cells.
The difficulty is entirely in the word "confidently."

## Why the most abundant soup genes are the wrong choice

The obvious move is to use whatever is most abundant in the soup, since those
give the strongest signal. Look again at the top of the soup profile above:
`MALAT1`, `B2M`, `TMSB4X`, `EEF1A1`, and a dozen ribosomal protein genes. Those
are the worst possible choices, because every cell expresses them. There is no
set of cells for which we can say the true expression is zero, so any
"contamination" we measure in them is mostly real biology.

`RPL10` makes the point concretely: it sits ninth in the soup profile, at 0.81%
of all background counts. Abundant — and useless for our purpose, because it is
abundant *everywhere*. A marker map for it looks busy, but that signal is genuine
ribosomal expression in every cluster, not contamination.

```{r rpl10_map, fig.width=7, fig.height=5}
gg <- SoupX::plotMarkerMap(sc, "RPL10")
plot(gg)
```

The useful genes are the opposite kind: **highly expressed in one cell type and
genuinely absent everywhere else**. `LYZ`, the monocyte marker from the first
tutorial, is that sort of gene. So is `IGKC`, an antibody gene made only by B
cells, and so is haemoglobin (`HBB`, `HBA2`) in tissue where red cells are the
only source.

## Letting the data suggest candidates

`plotMarkerDistribution()` gives a survey. For each candidate gene it shows the
distribution across cells of the ratio of observed counts to what pure soup would
produce, with a red line marking the contamination fraction you would infer from
that gene alone.

```{r marker_distribution, fig.width=9, fig.height=5}
SoupX::plotMarkerDistribution(sc)
```

> **Read this plot, do not automate it.** The SoupX authors are emphatic on this
> point: it is a heuristic to develop your intuition, and if you simply take the
> top N genes from it and estimate contamination with them, **you will
> over-estimate**. That is why the function returns a plot and not a gene list.

For a systematic candidate set we use `quickMarkers()`, which finds the genes
most specific to each cluster. Specificity is scored by tf-idf, the same measure
used to find words that characterize a document.

```{r quick_markers}
marker_genes <- SoupX::quickMarkers(sc$toc, sc$metaData$clusters)
head(marker_genes[, c("gene", "cluster", "tfidf", "qval")])
```

The top hits for cluster 1 — `LEF1`, `CCR7`, `MAL`, `FLT3LG` — are naive T-cell
genes, with q-values that leave no doubt. `quickMarkers()` has found real biology
without being told any, which is the whole reason it is usable on a dataset you
have never seen.

We reshape the result into a list: one entry per cluster, containing that
cluster's marker genes.

```{r cluster_genes}
cluster_genes <- split(marker_genes$gene, marker_genes$cluster)
lengths(cluster_genes)
```

Ten genes per cluster, because `quickMarkers()` defaults to `N = 10` — it returns
the top ten markers per cluster, not every gene that passes significance. Raising
`N` gives more genes to estimate from but admits progressively less specific ones,
which is a trade-off rather than a free improvement.

Note that we split on `marker_genes$gene`, the **column** holding the gene names.
`quickMarkers()` leaves the row names as `"1"`, `"2"`, `"3"`, … — the same trap as
with the clusters CSV above, and it bites the same way: `split()` on row names
returns lists of row numbers, and SoupX would then look for genes literally named
`"1"` and find nothing.

The reason `nonExpressedGeneList` is a *list* of gene sets rather than one flat
vector is that genes are grouped by biology. Immunoglobulin genes rise and fall
together; haemoglobin genes rise and fall together. SoupX uses or excludes each
group as a unit.

# Choosing which cells to estimate from

Having picked gene sets, we now need, for each set, the cells that can be assumed
not to express it. A B-cell marker tells us nothing about contamination *in B
cells* — there the expression is real.

```{r estimate_non_expressing}
useToEst <- SoupX::estimateNonExpressingCells(sc, nonExpressedGeneList = cluster_genes)
dim(useToEst)
```

The result is 4,340 × 8 — one row per cell, one column per gene set — holding
TRUE/FALSE: may this cell be used to estimate contamination with this gene set?

SoupX is deliberately conservative here. Because we supplied clusters, it works
at cluster level: if *any* cell in a cluster shows convincing evidence of
expressing a gene set, the entire cluster is excluded for that set. Excluding
real expressers matters much more than losing some cells, since a cell that
genuinely expresses the gene would inflate `rho` and cause over-correction.

We can look at the decision for a couple of gene sets. Cells outlined in green
are the ones being used for estimation; the fill colour is the observed-to-soup
ratio.

```{r plot_use_to_est, fig.width=7, fig.height=5}
SoupX::plotMarkerMap(sc, geneSet = cluster_genes[[1]], useToEst = useToEst[, 1])
SoupX::plotMarkerMap(sc, geneSet = cluster_genes[[5]], useToEst = useToEst[, 5])
```

The first set is cluster 1's markers — the naive T-cell genes `LEF1`, `CCR7`,
`MAL` and company from the table above. In each map, the cluster the markers came
from should be the region that is *excluded*: reddish, because expression there
is genuine and far above soup level, and not outlined, because those cells are
useless for measuring contamination. Everything else is available for estimation.
This is the step the automatic method performs invisibly, and it is where a
mistake would do the most damage.

# Calculating the contamination fraction

All the judgement has now been made. The arithmetic is one call.

```{r calc_contamination, message=TRUE}
sc <- SoupX::calculateContaminationFraction(sc, cluster_genes, useToEst = useToEst)
head(sc$metaData)
```

**9.78%** — nearly ten times the `rho = 0.01` of the first tutorial. Roughly one
count in ten in this dataset came from the soup rather than from the cell it was
assigned to. Now the correction has something substantial to do.

Unlike `autoEstCont()`, this function also records `rhoLow` and `rhoHigh`, a
confidence interval on the estimate. Here it runs from 0.0953 to 0.1004 — a
spread of half a percentage point, which is tight. That tells us the gene sets
carried plenty of information and the estimate is well determined. A wide
interval, or one running down towards 0, would be a signal to revisit the gene
and cell choices above rather than to accept the number.

> **A number that does not match the vignette, and why.** The SoupX vignette gets
> 7.69% on this same dataset; we get 9.78%. The difference is the gene sets. The
> vignette estimates from immunoglobulin genes only — a set its authors were
> confident about on biological grounds — whereas we used the top markers of all
> eight clusters. More gene sets means more cells contributing, but also admits
> genes whose "non-expression" is less certain, which pushes the estimate up.
> Neither number is the true value; both are estimates from different, defensible
> choices. This is worth sitting with, because it is the honest situation you are
> in with your own data: the contamination fraction is not read off the data, it
> is *inferred from assumptions you make*, and reasonable assumptions differ by a
> couple of percentage points. Recall from the first tutorial that SoupX is
> forgiving of exactly this kind of error.

## How this compares to the automatic route

It is worth being clear about what actually differed. `autoEstCont()` performs
these same steps: it calls `quickMarkers()`, applies `estimateNonExpressingCells()`,
and computes `rho` from many gene–cluster pairs. The differences are that it
filters the marker list more aggressively (by tf-idf and by abundance in the
soup), it combines the many per-gene estimates by taking the mode of their
posterior rather than a single pooled calculation, and it constrains the answer
to `contaminationRange`.

So this is not a different method so much as the same method with the decisions
exposed. The value of doing it by hand is that you can substitute your own
biology at the point where it matters. If you know your tissue contains red blood
cells, `list(HB = c("HBB", "HBA2"))` is a better gene set than anything
`quickMarkers()` will find for you.

# Correcting the counts

```{r adjust}
out <- SoupX::adjustCounts(sc)
```

# What actually changed

As before, the fraction of expressing cells driven to zero for each gene:

```{r most_zeroed}
cnt_soggy <- Matrix::rowSums(sc$toc > 0)
cnt_strained <- Matrix::rowSums(out > 0)
most_zeroed <- tail(sort((cnt_soggy - cnt_strained) / cnt_soggy), n = 10)
most_zeroed
```

Compare with the first tutorial, where the largest value was 0.14. Here the list
runs from 0.46 to **0.63**: for `S100A9`, 63% of the cells that had a non-zero
count now have none. Nearly ten times the soup produced a categorically different
correction, not merely a larger one.

The genes themselves are the story. Every one is a high-expression marker of a
single population:

- `S100A9`, `S100A8`, `LYZ`, `CST3` — monocytes and myeloid cells.
- `IGLC2`, `IGKC`, `IGLC3` — immunoglobulin, made only by B cells.
- `GNLY`, `NKG7` — cytotoxic granule genes from NK and effector T cells.
- `PPBP` — platelets.

This is the same pattern as the first tutorial, where the equivalent list was led
by platelet and myeloid genes, but far more pronounced. Note that `LYZ` (0.577)
and `IGKC` (0.575) — the two genes we look at directly below — are themselves in
this top ten, out of all 33,694 genes. They are not incidental examples; they are
among the genes the correction touches most.

And the genes that changed quantitatively:

```{r quantitative_change}
tail(sort(Matrix::rowSums(sc$toc > out) / Matrix::rowSums(sc$toc > 0)), n = 20)
```

The mitochondrial genes dominate, as they did in the first tutorial: `MT-ND1`
through `MT-CYB`, every one at a ratio of 1, meaning every cell expressing them
lost counts. Mitochondrial transcripts are over-represented in the background
relative to cells, so this is the expected result and it carries the same warning
as before — **ambient RNA inflates the mitochondrial fraction you filter on
during quality control**, so some cells that look low-quality are simply soupy.

The same caveat as the first tutorial also applies to the rest of the list.
`BX004987.4`, `AC011043.1`, `AL592183.1` and the other accession-numbered loci
are detected in very few cells, so the denominator of the ratio is tiny and
reaching exactly 1.0 takes almost nothing. Check the denominator before reading
meaning into the top of a sorted ratio.

## The correction, seen directly

```{r change_map_lyz, fig.width=7, fig.height=5}
SoupX::plotChangeMap(sc, out, "LYZ")
```

```{r change_map_igkc, fig.width=7, fig.height=5}
SoupX::plotChangeMap(sc, out, "IGKC")
```

`LYZ` should lose its scatter outside the monocyte cluster and `IGKC` outside the
B cells, while both keep their counts where the expression is real. We know from
the table above that this is a large effect — each gene lost its counts entirely
in about 57% of the cells that had any — so on this dirtier dataset the change
should be far more visible than the near-invisible correction in the first
tutorial.

Contrast that with a ubiquitously expressed gene:

```{r change_map_rpl10, fig.width=7, fig.height=5}
SoupX::plotChangeMap(sc, out, "RPL10")
```

`RPL10` is abundant in the soup, so counts are removed from it too — but they are
removed roughly evenly everywhere, because every cell genuinely expresses it.
Nothing about the interpretation of `RPL10` changes. This is the visual version
of the earlier point: correction reshapes the genes whose expression is
concentrated in one cell type, and merely rescales the ones that are not.

# Using the cleaned matrix downstream

```{r seurat, eval=FALSE}
pbmc <- Seurat::CreateSeuratObject(counts = out)
```

# What we glossed over

- **We still took CellRanger's clusters as given.** We attached them by hand
  rather than having them attached for us, but we did not produce or evaluate
  them. Clustering is Week 3. The good news is that SoupX is not very sensitive
  to the clustering used — Seurat defaults give similar results — because the
  clusters only need to group cells well enough to tell which ones express a
  gene set.
- **The gene sets came from the data, not from biology.** We used
  `quickMarkers()` because it generalizes to a dataset you know nothing about.
  When you *do* know the biology, naming the genes yourself is better — and as
  the 9.78%-versus-7.69% discrepancy above shows, the choice moves the answer by
  a couple of percentage points. For a tissue with red cell contamination, for
  example: `list(HB = c("HBB", "HBA2"), IG = c("IGKC", "IGHG1", "IGHM"))`.
- **We never cross-checked with `autoEstCont()`.** This dataset has clusters
  attached now, so the automatic estimator would run on it. Comparing the two
  numbers costs one line and is a genuinely good habit.
- **We accepted the default FDR thresholds** in `estimateNonExpressingCells()`,
  which control how much evidence of expression is needed to exclude a cluster.
- **`rho` is still one number for the whole channel.** Both routes assume a
  constant contamination fraction across cells within a channel.
- **We did not verify that the correction helped.** The honest test is downstream:
  cluster the corrected and uncorrected matrices and see whether marker genes
  become more cell-type specific. That is a Week 3 exercise, and a good one.

# References

- Young MD, Behjati S. *SoupX removes ambient RNA contamination from droplet-based
  single-cell RNA sequencing data.* GigaScience 9(12), 2020.
  [doi:10.1093/gigascience/giaa151](https://doi.org/10.1093/gigascience/giaa151)
- PBMC 4k dataset (10x Genomics, CellRanger 2.1.0). The files used above are
  downloaded directly from
  `https://cf.10xgenomics.com/samples/cell-exp/2.1.0/pbmc4k/`; the dataset's
  documentation page is
  [support.10xgenomics.com/single-cell-gene-expression/datasets/2.1.0/pbmc4k](https://support.10xgenomics.com/single-cell-gene-expression/datasets/2.1.0/pbmc4k).
