---
title: "Week 3: Batch effects and integration in Seurat v5"
author: "Kevin Z. Lin"
date: "2026-08-09"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial was made from the Seurat v5 vignette *Integrative analysis in
Seurat v5*,
[https://satijalab.org/seurat/articles/seurat5_integration](https://satijalab.org/seurat/articles/seurat5_integration),
and the Azimuth mapping tutorial,
[https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html](https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html).

Week 1 ran the whole pipeline on a single sample and pretended that was the
normal case. It is not. Almost nobody has one sample. You have four donors, or
two 10x runs done three weeks apart, or a pilot on one chemistry and the rest on
another. And the moment you have more than one, a new failure mode appears: **the
cells group by where they came from rather than by what they are.**

That is a batch effect, and it is not subtle. Run the Week 1 pipeline unchanged
on a multi-sample dataset and you will often get one group per sample per cell
type — a UMAP that looks like a technical inventory of your experiment rather
than a picture of your biology.

The fix is called **integration**, and the single most important thing to
understand about it is what it does *not* do:

> **Integration does not change your counts.** It produces a new
> low-dimensional embedding — a replacement for the PCA — in which cells of the
> same type from different batches sit on top of each other. Everything that
> consumes a reduction — UMAP, clustering, neighbour graphs — can then use it.
> But when you later test genes for differential expression, you go back to the
> original counts. Nothing in this tutorial edits a single count.

The plan:

| Stage | What it produces | Function |
|---|---|---|
| 0. Label the cells against a reference | A cell type per cell, fixed before any integration happens | `RunAzimuth()` |
| 1. Declare the batches | One layer per batch inside the RNA assay | `split()` |
| 2. Run the Week 1 pipeline | Normalized data, variable genes, a PCA | `NormalizeData()` … `RunPCA()` |
| 3. Look at the damage | A UMAP in which cells group by technology | `RunUMAP()`, `DimPlot()` |
| 4. Integrate | A new reduction, `integrated.cca`, in which they do not | `IntegrateLayers()` |
| 5. Look again | The same cells, the same genes, batches mixed | `RunUMAP()`, `DimPlot()` |
| 6. Put the layers back together | One `counts` and one `data` layer, ready for testing | `JoinLayers()` |

Stage 0 is there for a specific reason. To judge whether integration worked, you
need a read on each cell's identity that does not itself depend on the
integration. Anything computed *from* the integrated embedding — clusters above
all — is circular evidence that it succeeded. Azimuth gives us an independent
label per cell, assigned against an external reference, before any of the
integration happens, and we never update it afterwards.

That is also why this tutorial never clusters. Integration produces a reduction;
whether it worked is a question about that reduction, and two fixed labels —
which batch a cell came from and which cell type it is — answer it without
introducing a clustering algorithm and its resolution parameter as a confound.

> **Where this sits in the workshop.** The middle of this tutorial — normalize,
> pick variable genes, scale, PCA — is exactly the Week 3 material, run
> here on a dataset whose batch structure is impossible to ignore. The companion
> Week 3 tutorial takes the normalization step on its own and replaces this
> log-normalize / variable-genes / scale sequence with `SCTransform`. Batch
> correction runs across two sessions: this is part one, and **Week 4** returns
> to it — comparing integration methods and testing genes once the batches are
> aligned. Bulk deconvolution is **Week 5**.

# Setting up

`Seurat` is on CRAN. `SeuratData` (which distributes the demonstration datasets)
and `Azimuth` (which does the reference mapping) are not, so they install from
GitHub with `devtools`. You only need to do this once, which is why the lines are
commented out.

```{r install}
# install.packages("devtools")
# devtools::install_github("satijalab/seurat-data")
# devtools::install_github("satijalab/azimuth")
```

```{r libraries}
library(Seurat)
library(SeuratData)
library(SeuratObject)
library(Azimuth)
library(patchwork)
library(ggplot2)
```

```{r clean_env}
rm(list = ls())
```

Integration ships pieces of your data to parallel workers, and Seurat's default
size limit for doing so is smaller than this dataset. Raising it up front avoids
an error two hundred lines from now that says nothing about batches:

```{r future_globals}
options(future.globals.maxSize = 1e10)
```

> **If `RunAzimuth()` fails with a `stringi` error.** A `lazy-load database
> '.../stringi/R/stringi.rdb' is corrupt` message has nothing to do with your
> data or with Azimuth. It means the `stringi` package was updated underneath a
> running R session. Restart R and run the script again.

# The data

We use **pbmcsca**, the dataset from Ding *et al.* (2020), who took peripheral
blood mononuclear cells from two donors and profiled them with seven different
single-cell technologies. It is a deliberately extreme case: the technologies
differ in chemistry, in capture efficiency, in read depth, and in whether they
count molecules or reads at all.

That extremity is the point. Your own batch effect will be milder than this one,
but it will be the same *kind* of thing, and a mild version of a visible problem
is easier to learn from than a mild version of an invisible one.

```{r install_data}
# SeuratData::InstallData("pbmcsca")
```

```{r load_data}
obj <- SeuratData::LoadData("pbmcsca")
obj
```

```{r methods_table}
table(obj$Method)
```

Nine batches, not seven: 10x Chromium v2 was run three separate times, and those
runs are correctly treated as separate batches even though the chemistry is
identical. **A batch is a unit of technical variation, not a brand name.**

The metadata also carries the authors' own cell type calls, which we will use
later as a second opinion:

```{r metadata_head}
head(obj@meta.data)
```

## A filter that is itself a batch effect

The vignette opens by dropping cells with fewer than 1,000 detected genes. This
looks like the routine quality filter from Week 1, and it is worth pausing on
because it is not routine at all here.

```{r subset}
obj <- subset(obj, subset = nFeature_RNA > 1000)
obj
```

```{r methods_table_after}
table(obj$Method)
```

Compare that with the table before the filter. Smart-seq2 kept 512 of its 526
cells; inDrops kept 545 of 6,584. **One global threshold retained 97% of one
technology and 8% of another** — not because inDrops captured worse cells, but
because a plate-based full-length protocol and a high-throughput droplet protocol
detect genes at completely different rates.

> **The general lesson, which matters more than this dataset.** A single QC
> cutoff applied across batches silently reweights your experiment. If your
> batches differ in depth — different chemistries, different sequencing runs,
> fresh versus frozen — a global `nFeature_RNA` floor removes cells
> disproportionately from the shallow batches, and you will never see it in the
> filtered object. Tabulate your QC threshold against your batch variable before
> you trust it.

The filter takes 31,021 cells down to 10,434. We are keeping it, because it is
what the vignette does and because the surviving cells still show the batch
effect clearly — but it is a choice, not a default.

## The batch effect, visible in five numbers

Before any modelling, look at the raw matrix.

```{r layer_data}
count_mat <- SeuratObject::LayerData(obj, assay = "RNA", layer = "counts")
dim(count_mat)
count_mat[1:5, 1:5]
```

Those five columns are Smart-seq2 cells, and `DPM1` has a count of 366 in the
first one. A housekeeping gene in a 10x cell would show a count in the single
digits. Nothing is wrong: Smart-seq2 counts *reads*, while the droplet methods
count *unique molecules*, so the two are on scales that are not comparable at
all. Per batch:

```{r depth_by_method}
tapply(obj$nCount_RNA, obj$Method, median)
```

A median library of 331,917.5 for Smart-seq2 against 3,098 for 10x Chromium v2 —
a factor of **107** between two sets of cells in the same object. (The `.5`
endings are just medians over an even number of cells.)

That is the batch effect. It is sitting in the count matrix before you do
anything, and no amount of log-normalization makes two protocols that measure
different quantities comparable. Notice also that the other eight batches span
only 2,572 to 6,083, so the spread here is really "Smart-seq2 versus everything
else" — a reminder that batches are rarely equally different from one another,
and that one outlying batch can dominate what integration ends up correcting.

# Labelling the cells with Azimuth

`RunAzimuth()` maps our cells onto an annotated reference and copies the
reference's labels across. The reference here is `pbmcref`, the Satija lab's
multimodal PBMC atlas. Under the hood it normalizes our query with the
reference's own SCTransform model, finds **transfer anchors** — pairs of cells,
one in the query and one in the reference, that look like mutual nearest
neighbours — and uses those anchors to carry labels over.

```{r run_azimuth}
obj <- Azimuth::RunAzimuth(obj, reference = "pbmcref")
obj
```

The object grew from one assay to four. The three new ones are named
`prediction.score.celltype.l1`, `.l2` and `.l3`, and they are the first surprise:

> **The `prediction.score.*` assays do not contain expression.** An "assay" in
> Seurat is just a features-by-cells matrix, and Azimuth reuses that container to
> store, for each cell, the probability assigned to *every* candidate cell type.
> The rows are cell types, not genes. Running `FindVariableFeatures()` or a
> differential expression test on one of these by accident is a real hazard —
> check `DefaultAssay(obj)` before any command that does not name its assay.

The reference is annotated at three levels of resolution, so we get three sets of
labels — from eight broad classes at level 1 down to fine subtypes at level 3:

```{r azimuth_labels}
table(obj$predicted.celltype.l1)
length(unique(obj$predicted.celltype.l2))
```

Two reductions also appeared, `integrated_dr` and `ref.umap`. Those are the
query cells projected into the *reference's* coordinate system. They are useful,
but they are not what we integrate on — we build our own embedding below.

## Reading the confidence, not just the label

Every cell got a label. That is not a compliment to the data; Azimuth has no
"unknown" category, so the nearest reference cell type always wins, however far
away it is. Two columns tell you how much to believe it:

```{r azimuth_scores}
quantile(obj$predicted.celltype.l2.score)
quantile(obj$mapping.score)
```

The **prediction score** is how strongly the anchors agreed on this particular
label; the **mapping score** is a different question — whether the cell's local
neighbourhood is represented in the reference at all. A cell can score highly on
the first and poorly on the second, and that combination is the interesting one:
confidently assigned to a type that does not really describe it.

The median prediction score is 0.918, which is high, but the minimum is 0.219.
Those low-scoring cells still carry a definite-looking string in
`predicted.celltype.l2`, and nothing downstream will remind you that it was a
coin flip.

## A second opinion

This dataset is unusual in shipping the original authors' own annotations, so we
can compare two independent labelings of the same cells:

```{r label_agreement}
table(obj$predicted.celltype.l1, obj$CellType)
```

Read the diagonal blocks first: 1,489 of the 1,525 published B cells are called
`B`, and 1,493 of 1,557 CD14+ monocytes are called `Mono`. The coarse structure
is not in dispute.

Now read off the diagonal, where it gets honest. The published "Cytotoxic T
cell" class splits across `CD8 T` (1,846), `other T` (569), `CD4 T` (201) and
`NK` (167). Neither labeling is wrong — they were made against different
references with different granularity, and the CD8/NK/γδ boundary is genuinely
where T cell annotation is hardest.

> **The transferable point for your own data.** Cell type labels are estimates
> with error bars, not facts read off the cells. Two competent analyses of the
> same PBMCs — the easiest, most heavily annotated tissue there is — place about
> a third of the T cells in different classes. If a conclusion in your paper
> depends on a boundary between two adjacent labels, that boundary needs
> defending.

> **`pbmcref` is a PBMC reference.** Everyone in this room is working on brain,
> and mapping nuclei onto a blood atlas will produce confident, meaningless
> labels. Azimuth ships human motor cortex among its references, and MapMyCells
> covers the Allen brain taxonomies — that is Week 4's material. What generalizes
> from this section is the *procedure*, not the reference.

# Declaring the batches

Here is the idea that organizes all of Seurat v5. In v4, nine batches meant nine
Seurat objects that you merged and tracked by hand. In v5 there is one object,
and the assay inside it holds **layers** — one `counts` and one `data` layer per
batch.

```{r split_layers}
obj[["RNA"]] <- split(obj[["RNA"]], f = obj$Method)
obj
```

Eighteen layers now: nine `counts` and nine `data`, named after the values of
`Method`. The cells did not move and nothing was copied apart from the split
itself.

> **`f =` is the most consequential argument in this tutorial.** It defines what
> you are calling a batch, and therefore what integration will treat as noise to
> be removed. Split by 10x run and you remove run-to-run variation. Split by
> donor and you remove donor-to-donor variation — including, if you are not
> careful, the difference between your cases and your controls. More on this
> below; it is the one part of this tutorial that can invalidate a paper.

# The Week 1 pipeline, unchanged

Nothing here is new except what the layers do to it.

```{r normalize}
obj <- Seurat::NormalizeData(obj)
obj <- Seurat::FindVariableFeatures(obj)
obj <- Seurat::ScaleData(obj)
```

Because the assay is split, `NormalizeData()` and `FindVariableFeatures()` run
**once per layer** rather than once overall. That matters: variable genes chosen
across the pooled data would be dominated by genes that differ between
technologies, which are exactly the genes you do not want driving the structure.
Seurat picks the variable genes per batch and then takes a consensus set.

```{r variable_features}
length(Seurat::VariableFeatures(obj))
head(Seurat::VariableFeatures(obj), 20)
```

Two thousand genes, led by `LYZ`, `S100A9`, `CD74` and the MHC class II genes —
a sensible list of things that differ between monocytes, B cells and T cells.

```{r pca}
set.seed(10)
obj <- Seurat::RunPCA(obj, verbose = FALSE, seed.use = 10)
```

`RunPCA()` draws random starting vectors, so it gets a seed. Seurat's own
`seed.use` argument is set as well as the ambient stream, because several Seurat
functions call `set.seed()` internally just before their solver — setting only
the outer seed looks reproducible without being reproducible.

```{r umap_unintegrated}
set.seed(10)
obj <- Seurat::RunUMAP(obj,
                       dims = 1:30,
                       reduction = "pca",
                       reduction.name = "umap.unintegrated",
                       seed.use = 10)
```

`reduction.name` matters here: the default would be `umap`, and the second UMAP
we build after integration would silently replace this one. We want both, to
compare.

# Looking at the damage

```{r dimplot_unintegrated_method, fig.width=8, fig.height=5}
Seurat::DimPlot(obj, reduction = "umap.unintegrated", group.by = "Method")
```

```{r dimplot_unintegrated_celltype, fig.width=9, fig.height=5}
Seurat::DimPlot(obj, reduction = "umap.unintegrated",
                group.by = "predicted.celltype.l2")
```

Hold these two panels side by side. In the first, the islands are almost
uniformly single-coloured: Drop-seq occupies its own territory, Smart-seq2 and
Seq-Well theirs, and only the three 10x runs partially overlap each other. In the
second, the cell type labels *do* look coherent within each island — but the same
label keeps reappearing in islands far apart. `CD14 Mono` shows up as several
separate blobs on the right, `B naive` as three or four along the top.

That is the signature of an unintegrated multi-batch dataset: not one island per
cell type, but one island per cell type **per batch**.

Before integrating, there is one prerequisite worth checking. Anchor-based
methods assume the batches contain **the same cell types** — that is what makes
"this cell in batch A matches that cell in batch B" a meaningful statement. The
Azimuth labels let us check that directly, without any clustering:

```{r celltype_by_method}
table(obj$predicted.celltype.l1, obj$Method)
```

No cell type is missing from any batch, which is the condition integration needs.
But look at how unevenly they are represented: **inDrops is 90% monocytes, while
Drop-seq is 3%**. Seq-Well is 54% monocytes; CEL-Seq2 and Smart-seq2 are more
than half T cells.

That is a second kind of batch effect, and integration does not fix it. Aligning
the batches makes a monocyte from one protocol land on a monocyte from another;
it does not change the fact that one protocol captured 490 monocytes and another
captured 58. Any comparison of cell type *proportions* across these batches is
measuring capture chemistry, not biology.

> **If a batch genuinely lacks a cell type, integration will invent one.** The
> method's job is to align populations across batches; handed a batch with no
> monocytes, it will still find the "best" match for the other batches'
> monocytes and pull something onto them. Check this table before you integrate,
> not after you have a UMAP you like.

# Integrating

`IntegrateLayers()` takes the per-batch layers and the unintegrated PCA, and
returns a new reduction in which the batches are aligned.

```{r integrate}
obj <- Seurat::IntegrateLayers(object = obj,
                               method = Seurat::CCAIntegration,
                               orig.reduction = "pca",
                               new.reduction = "integrated.cca",
                               verbose = FALSE)
obj
```

This takes a couple of minutes on a laptop — the slowest step in the tutorial.

`CCAIntegration` is Seurat's anchor-based method. Canonical correlation analysis
finds directions of variation that are *shared* between two datasets rather than
maximal within either one; cells that are mutual nearest neighbours in that
shared space become **anchors**, and the anchors define the correction applied to
everything else. The intuition worth keeping is that anchors are the method's
guess at "this cell in batch A and this cell in batch B are the same kind of
cell", and everything the method does downstream is only as good as that guess.

Four other methods are available through the same one-line interface, and
switching between them is a change of one argument:

| `method =` | What it is |
|---|---|
| `CCAIntegration` | Anchor-based CCA. Strong correction; what we use here |
| `RPCAIntegration` | Anchor-based, reciprocal PCA. Faster and more conservative |
| `HarmonyIntegration` | Iterative clustering and correction in PCA space |
| `FastMNNIntegration` | Mutual nearest neighbours |
| `scVIIntegration` | Variational autoencoder; needs Python and `scvi-tools` |

Which to use is a real question with no universal answer, and it is Week 4's
material. The relevant fact for today is that trying a second one costs one line,
so "we used CCA because it was in the tutorial" is not a defensible methods
section when the alternative was that cheap.

```{r umap_integrated}
set.seed(10)
obj <- Seurat::RunUMAP(obj,
                       reduction = "integrated.cca",
                       dims = 1:30,
                       reduction.name = "umap.cca",
                       seed.use = 10)
```

Note that `RunUMAP()` is pointed at `integrated.cca` rather than `pca`. That is
the entire behavioural difference between the integrated and unintegrated
analyses: same object, same genes, same counts, different reduction.

# Looking again

```{r dimplot_integrated_method, fig.width=8, fig.height=5}
Seurat::DimPlot(obj, reduction = "umap.cca", group.by = "Method")
```

```{r dimplot_integrated_celltype, fig.width=9, fig.height=5}
Seurat::DimPlot(obj, reduction = "umap.cca",
                group.by = "predicted.celltype.l2")
```

Compare the first panel with its counterpart above. The single-coloured islands
are gone; every island is now a speckled mixture of all nine technologies. One
region at the bottom is still visibly blue, which we come back to in a moment.

The second panel is the one that matters, though, and it is why we ran Azimuth
first. The Azimuth labels were assigned before integration and never updated, so
the fact that the islands are now coherent in those labels — B cells along the
top, monocytes along the bottom, T cells through the middle, each as **one**
territory rather than several — is not something integration manufactured. It is
the check that the mixing removed technology rather than biology.

## What did not get fixed

Integration is not erasure, and it is worth looking for what survived rather
than declaring victory. One region near the bottom of the integrated UMAP is
still visibly blue — a patch where inDrops cells sit largely on their own instead
of mixing. Reading it against the cell type panel, it sits inside the monocyte
territory, which is exactly where the table above said to expect trouble: inDrops
contributed 490 monocytes and very little else, so that population has far more
inDrops cells to anchor to each other than to anything from the other batches.

A residual batch-specific patch is not automatically a failure. It can mean the
correction was incomplete, or it can mean that one protocol genuinely captured a
state the others missed — which, given that these nine batches differ in capture
chemistry and lysis, is entirely plausible. The two explanations look identical
in a UMAP. Telling them apart needs marker genes, not more integration.

# Putting the layers back

Splitting the assay was for integration's benefit. Almost everything downstream —
differential expression above all — expects a single `counts` layer and a single
`data` layer rather than nine of each.

```{r join_layers}
obj <- SeuratObject::JoinLayers(obj)
obj
```

Eighteen layers collapse back to three: `counts`, `data`, and the `scale.data`
that `ScaleData()` produced (it was never split — scaling ran on the joined
matrix all along).

> **`JoinLayers()` lives in `SeuratObject`, not `Seurat`.** Writing
> `Seurat::JoinLayers(obj)` gives `'JoinLayers' is not an exported object from
> 'namespace:Seurat'`. Because both packages are attached by `library(Seurat)`,
> the unqualified `JoinLayers(obj)` works fine, which is why the mistake survives
> until someone adds namespaces. `Seurat::` is not a safe blanket prefix for
> everything in the Seurat ecosystem.

If you want to integrate again — a different method, a different `f =` — you can
re-split at any time. The split is bookkeeping, not a transformation.

Finally, the prediction score matrices are still there, and worth seeing once so
that the shape is not a surprise later:

```{r prediction_scores}
score_mat <- SeuratObject::LayerData(obj,
                                     assay = "prediction.score.celltype.l1")
dim(score_mat)
rownames(score_mat)
round(score_mat[1:5, 1:5], 3)
```

Eight rows for the eight level-1 cell types, one column per cell, and each column
sums to one. `predicted.celltype.l1` is just the row name of the largest entry in
each column; this matrix is where the rest of the distribution went.

# The two ways to get this badly wrong

Everything above worked because `Method` is very nearly a purely technical
variable, and because nobody cares which of two healthy donors a PBMC came from.
Two mistakes break that, and both are common.

## Confounding batch with your biology

Suppose you have twelve donors, six with Alzheimer's disease and six controls,
and — as happens constantly — the cases were processed in one batch and the
controls in another. Integrate on batch and you will remove the disease effect,
because in that design the disease effect *is* the batch effect. The UMAP will
look beautiful. The differential expression will find nothing, or worse, find
something.

No integration method can separate two variables that are perfectly confounded;
this is a property of the design, not a deficiency of the software. What you can
do:

- **Check first.** `table(obj$batch, obj$condition)` takes one second, and
  answers whether you have a problem before you spend a day integrating.
- **Balance at the bench.** Split conditions across batches when the samples are
  being processed. This is the only real fix, and it happens before any of us see
  the data.
- **When it is partial**, integrate on the technical variable and keep the
  biological one out of the split. With four batches each containing both cases
  and controls, integrating on batch is safe.
- **Consider not integrating.** Keeping the batch variable as a covariate in a
  differential expression model — or moving to pseudobulk, which sidesteps much
  of this — is often the more defensible route than correcting the embedding.

Run that check on this very dataset before assuming we got away with it:

```{r confounding_check}
table(obj$Method, obj$Experiment)
```

`Experiment` is the donor. Four of the nine batches — 10x Chromium v2, v2 A,
v2 B and v3 — contain cells from **one donor only**, and Seq-Well is 337 to 13.
So `Method` and donor are *partially* confounded here, and some of what
integration removed was donor difference rather than protocol difference.

That is survivable in this dataset for two reasons: five batches do span both
donors, which is enough to anchor the alignment, and donor identity is not the
thing anyone is trying to measure in a methods-comparison paper. Neither reason
would hold if the two donors were a case and a control.

> **Do not skip this table on your own data.** It is one line, it runs before
> anything expensive, and it is the difference between removing a nuisance and
> removing your result. We ran it here and found partial confounding in a
> dataset specifically designed to be clean.

## Testing genes on integrated values

The integrated output is a reduction — thirty-odd dimensions with no gene names.
Some methods can also produce a corrected expression matrix, and it is tempting
to run differential expression on it.

Do not. Integration deliberately removed variation it decided was batch, using
anchors that assumed the batches contain the same cell types. Those assumptions
are fine for arranging cells in a low-dimensional space and inappropriate as
input to a statistical test, whose p-values assume the values are measurements
rather than reconstructions.

**Cluster on the integrated reduction; test on the original counts.** Week 4
takes this up in detail.

# What we glossed over

- **Why 30 dimensions?** Copied from the vignette, and consequential: the number
  of principal components fed to `IntegrateLayers()` and `RunUMAP()` sets how
  much structure either one can see. Choosing dimensionality — and clustering on
  top of the result — is the rest of **Week 3**.
- **Log-normalization versus `SCTransform`.** `IntegrateLayers()` accepts
  `normalization.method = "SCT"`, and the whole tutorial runs with SCTransform
  in place of `NormalizeData()` / `FindVariableFeatures()` / `ScaleData()`. The
  companion **Week 3** normalization tutorial covers what that substitution
  actually changes.
- **Which integration method, and how to tell.** We used CCA and never compared
  it with the four alternatives, and we never computed a real integration metric
  — kBET, LISI, silhouette by batch. We judged this one by eye, against labels
  fixed beforehand, which is honest but not quantitative. Note that any
  batch-mixing score alone rewards over-correction: a method that mashed
  everything into one blob would score perfectly, so mixing has to be read
  against a measure of preserved biology. **Week 4**.
- **Cell type labeling as a topic.** We ran one reference and read the scores.
  Choosing a reference, marker-based labeling, MapMyCells, and what to do when no
  reference fits your tissue are **Week 4**.
- **Differential expression after integration**, including pseudobulk and
  putting batch in the model rather than correcting it out — **Week 4**.
- **Ambient RNA and doublets.** We took `pbmcsca` as distributed and ran no
  Week 2 cleaning on it. On your own data, that cleaning comes first.
- **`mapping.score` went unused.** We computed it, quoted its quantiles, and
  never filtered on it. Cells that map poorly to the reference are exactly the
  cells whose labels should not be trusted in the evaluation above.

# References

- Ding J, Adiconis X, Simmons SK, *et al.* *Systematic comparative analysis of
  single-nucleus and single-cell RNA-sequencing methods.* Nature Biotechnology
  38:737–746, 2020.
  [doi:10.1038/s41587-020-0465-8](https://doi.org/10.1038/s41587-020-0465-8)
- Hao Y, Hao S, Andersen-Nissen E, *et al.* *Integrated analysis of multimodal
  single-cell data.* Cell 184:3573–3587, 2021.
  [doi:10.1016/j.cell.2021.04.048](https://doi.org/10.1016/j.cell.2021.04.048)
  — the source of the `pbmcref` reference and of the anchor-based framework.
- Seurat v5 vignette:
  [Integrative analysis in Seurat v5](https://satijalab.org/seurat/articles/seurat5_integration)
- Azimuth:
  [satijalab.github.io/azimuth](https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html)
